---
title: "Lecture 28: Antithetic variables"
author: "Ciaran Evans"
output: beamer_presentation
editor_options: 
  chunk_output_type: console
---

## Warmup: antithetic sampling

Work with your neighbor on the questions on the handout / course website:

[https://sta379-s25.github.io/practice_questions/pq_28_warmup.html](https://sta379-s25.github.io/practice_questions/pq_28_warmup.html)

Then we will discuss as a class


* Use Monte Carlo integration to approximate another integral
* Explore variability of two different estimators

## Warmup: antithetic sampling

```{r, include=F}
g <- function(x){
  exp(x)
}

n <- 1000
nsim <- 1000

theta_hat_mc <- rep(NA, nsim)
for(i in 1:nsim){
  u <- runif(n)
  theta_hat_mc[i] <- mean(g(u))
}

var(theta_hat_mc)
```


```{r}
theta_hat_as <- rep(NA, nsim)
for(i in 1:nsim){
  u <- runif(n/2)
  theta_hat_as[i] <- sum(g(u) + g(1-u))/n
}

(var(theta_hat_mc) - var(theta_hat_as))/var(theta_hat_mc)

u <- runif(10000)
cor(g(u), g(1-u))
```

\vspace{3cm}

## Antithetic variables

$$\theta = \mathbb{E}[g(U)] \hspace{1cm} U \sim Uniform(0, 1)$$
**Antithetic sampling:** 

* **Theorem:** If $g$ is *monotone*, then $Cor(g(U), \ g(1 - U)) \leq 0$

* Sample $U_1,...,U_{n/2} \overset{iid}{\sim} Uniform(0, 1)$

* $\widehat{\theta}_{AS} = \frac{1}{n} \sum \limits_{i=1}^{n/2} (g(U_i) + g(1 - U_i))$

* $Var(\widehat{\theta}_{AS}) = \dfrac{(1 + \rho)Var(g(U))}{n}$ where $\rho = Cor(g(U), g(1 - U))$


## Another example

Suppose we want to approximate the integral

$$\theta = \int \limits_1^\infty e^{-x} \frac{4}{x^5} dx = \mathbb{E}[e^{-X}]$$
where $X$ has pdf $f(x) = \dfrac{4}{x^5}$, $x > 1$.

**Simple Monte Carlo:**

* $X_1,...,X_n \overset{iid}{\sim} f$
* $\widehat{\theta}_{MC} = \frac{1}{n} \sum \limits_{i=1}^n e^{-X_i}$

**Question:** How do we sample $X_i \sim f$?

\vspace{3cm}

## Another example

Suppose we want to approximate the integral

$$\theta = \int \limits_1^\infty e^{-x} \frac{4}{x^5} dx = \mathbb{E}[e^{-X}]$$
where $X$ has pdf $f(x) = \dfrac{4}{x^5}$, $x > 1$.

**Simple Monte Carlo:**

* $U_1,...,U_n \overset{iid}{\sim} Uniform(0, 1)$
* $X_i = F^{-1}(U_i)$
* $\widehat{\theta}_{MC} = \frac{1}{n} \sum \limits_{i=1}^n e^{-X_i} = \frac{1}{n} \sum \limits_{i=1}^n e^{-F^{-1}(U_i)}$

**Question:** Can we use antithetic sampling here?


## Another example

Suppose we want to approximate the integral

$$\theta = \int \limits_1^\infty e^{-x} \frac{4}{x^5} dx = \mathbb{E}[e^{-X}]$$
where $X$ has pdf $f(x) = \dfrac{4}{x^5}$, $x > 1$.

\vspace{5cm}


## Antithetic sampling with inverse transform

Suppose we want to approximate

$$\theta = \mathbb{E}[g(X)] \hspace{1cm} X \sim f$$
If we can generate $X$ with the inverse transform method, then

$$\theta = \mathbb{E}[g(F^{-1}(U))] \hspace{1cm} U \sim Uniform(0, 1)$$

**Antithetic sampling:** 

* **Theorem:** If $g$ is *monotone*, then $Cor(g(F^{-1}(U)), \ g(F^{-1}(1 - U))) \leq 0$

* Sample $U_1,...,U_{n/2} \overset{iid}{\sim} Uniform(0, 1)$

* $\widehat{\theta}_{AS} = \frac{1}{n} \sum \limits_{i=1}^{n/2} (g(F^{-1}(U_i)) + g(F^{-1}(1 - U_i)))$


## Your turn

Try antithetic sampling with the inverse transform method:

[https://sta379-s25.github.io/practice_questions/pq_28.html](https://sta379-s25.github.io/practice_questions/pq_28.html)

* Start in class
* Welcome to work with a neighbor
* Solutions are posted on the course website



