---
title: "Lecture 19: Numerical Integration"
author: "Ciaran Evans"
output: beamer_presentation
editor_options: 
  chunk_output_type: console
---

## Where we are in the semester

Assignments:

* Due this week:
    * Project 1 (Wednesday -- if you use two extension days, would be due Friday)
    * HW 5 (Friday)
* Challenge 5 posted on course website
* Project 2 will be posted soon

Material:

* New topic: numerical integration
* Next topic: Monte Carlo integration


## Numerical integration

**Numerical integration:** Numerical approximations to definite integrals that are hard to solve in closed form.

\vspace{1cm}

**Question:** In statistics, when do integrals without a closed form solution arise?

\vspace{1cm}

## Numerical integration: motivation

* CDFs with intractable integrals: e.g.,

$$\int \limits_{-\infty}^t \frac{1}{\sqrt{2 \pi}} e^{-x^2/2} dx \hspace{1cm} \int \limits_{0}^t \frac{\Gamma(\alpha)\Gamma(\beta)}{\Gamma(\alpha + \beta)} x^{\alpha - 1}(1 - x)^{\beta - 1} dx$$

\vspace{1cm}

* Optimization problems involving integrals: e.g., mixed effects models,

$$\prod_i \int \left[ \phi(\gamma_i; 0, \sigma^2) \prod_j f(y_{ij} | \theta_{ij}) \right] d\gamma_i$$

## Numerical integration

Suppose I want to calculate $\int \limits_a^b f(x) dx$, but I can't get a closed form for the anti-derivative.

```{r, fig.width=3, fig.height=2.5, echo=F, message=F, warning=F}
library(tidyverse)
x <- seq(0, 1, 0.01)
y <- x^3 - 1.5*x^2 + x
#plot(x, y, type="l", ylab="f(x)", xaxt='n', yaxt='n', cex.lab=0.75)
data.frame(x = x, y = y) |>
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  theme_classic() +
  labs(y = "f(x)") +
  theme(axis.ticks.x=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y=element_blank())
```

**Question:** How can we approximate the integral without a closed form?

## Numerical integration: Riemann sum

Divide $[a, b]$ into $n$ subintervals:


```{r fig.width=3.5, fig.height=3, echo=F}
temp_f <- function(x){
  x^3 - 1.5*x^2 + x
}
data.frame(x = x, y = y) |>
  ggplot() +
  geom_line(aes(x = x, y = y)) +
  geom_segment(x = 0.1, y = 0, xend = 0.1, yend = temp_f(0.1), lty=2) +
  geom_segment(x = 0.2, y = 0, xend = 0.2, yend = temp_f(0.2), lty=2) +
  geom_segment(x = 0.3, y = 0, xend = 0.3, yend = temp_f(0.3), lty=2) +
  geom_segment(x = 0.4, y = 0, xend = 0.4, yend = temp_f(0.4), lty=2) +
  geom_segment(x = 0.5, y = 0, xend = 0.5, yend = temp_f(0.5), lty=2) +
  geom_segment(x = 0.6, y = 0, xend = 0.6, yend = temp_f(0.6), lty=2) +
  geom_segment(x = 0.7, y = 0, xend = 0.7, yend = temp_f(0.7), lty=2) +
  geom_segment(x = 0.8, y = 0, xend = 0.8, yend = temp_f(0.8), lty=2) +
  geom_segment(x = 0.9, y = 0, xend = 0.9, yend = temp_f(0.9), lty=2) +
  theme_classic() +
  labs(y = "f(x)") +
  theme(axis.ticks.x=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y=element_blank())
```


## Numerical integration: Riemann sum

Approximate $f(x)$ in each interval:


```{r fig.width=3.5, fig.height=3, echo=F}
data.frame(x = x, y = y) |>
  ggplot() +
  geom_line(aes(x = x, y = y)) +
  geom_segment(x = 0.1, y = 0, xend = 0.1, yend = temp_f(0.1), lty=2) +
  geom_segment(x = 0.1, y = temp_f(0.1), xend=0.2, yend=temp_f(0.1), lty=2) +
  geom_segment(x = 0.2, y = 0, xend = 0.2, yend = temp_f(0.2), lty=2) +
  geom_segment(x = 0.2, y = temp_f(0.2), xend=0.3, yend=temp_f(0.2), lty=2) +
  geom_segment(x = 0.3, y = 0, xend = 0.3, yend = temp_f(0.3), lty=2) +
  geom_segment(x = 0.3, y = temp_f(0.3), xend=0.4, yend=temp_f(0.3), lty=2) +
  geom_segment(x = 0.4, y = 0, xend = 0.4, yend = temp_f(0.4), lty=2) +
  geom_segment(x = 0.4, y = temp_f(0.4), xend=0.5, yend=temp_f(0.4), lty=2) +
  geom_segment(x = 0.5, y = 0, xend = 0.5, yend = temp_f(0.5), lty=2) +
  geom_segment(x = 0.5, y = temp_f(0.5), xend=0.6, yend=temp_f(0.5), lty=2) +
  geom_segment(x = 0.6, y = 0, xend = 0.6, yend = temp_f(0.6), lty=2) +
  geom_segment(x = 0.6, y = temp_f(0.6), xend=0.7, yend=temp_f(0.6), lty=2) +
  geom_segment(x = 0.7, y = 0, xend = 0.7, yend = temp_f(0.7), lty=2) +
  geom_segment(x = 0.7, y = temp_f(0.7), xend=0.8, yend=temp_f(0.7), lty=2) +
  geom_segment(x = 0.8, y = 0, xend = 0.8, yend = temp_f(0.8), lty=2) +
  geom_segment(x = 0.8, y = temp_f(0.8), xend=0.9, yend=temp_f(0.8), lty=2) +
  geom_segment(x = 0.9, y = 0, xend = 0.9, yend = temp_f(0.9), lty=2) +
  theme_classic() +
  labs(y = "f(x)") +
  theme(axis.ticks.x=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y=element_blank())
```


## Numerical integration: Riemann sum

Use the approximation in each interval to approximate the integral:


```{r fig.width=3.5, fig.height=3, echo=F}
data.frame(x = x, y = y) |>
  ggplot() +
  geom_line(aes(x = x, y = y)) +
  geom_segment(x = 0.1, y = 0, xend = 0.1, yend = temp_f(0.1), lty=2) +
  geom_segment(x = 0.1, y = temp_f(0.1), xend=0.2, yend=temp_f(0.1), lty=2) +
  geom_segment(x = 0.2, y = 0, xend = 0.2, yend = temp_f(0.2), lty=2) +
  geom_segment(x = 0.2, y = temp_f(0.2), xend=0.3, yend=temp_f(0.2), lty=2) +
  geom_segment(x = 0.3, y = 0, xend = 0.3, yend = temp_f(0.3), lty=2) +
  geom_segment(x = 0.3, y = temp_f(0.3), xend=0.4, yend=temp_f(0.3), lty=2) +
  geom_segment(x = 0.4, y = 0, xend = 0.4, yend = temp_f(0.4), lty=2) +
  geom_segment(x = 0.4, y = temp_f(0.4), xend=0.5, yend=temp_f(0.4), lty=2) +
  geom_segment(x = 0.5, y = 0, xend = 0.5, yend = temp_f(0.5), lty=2) +
  geom_segment(x = 0.5, y = temp_f(0.5), xend=0.6, yend=temp_f(0.5), lty=2) +
  geom_segment(x = 0.6, y = 0, xend = 0.6, yend = temp_f(0.6), lty=2) +
  geom_segment(x = 0.6, y = temp_f(0.6), xend=0.7, yend=temp_f(0.6), lty=2) +
  geom_segment(x = 0.7, y = 0, xend = 0.7, yend = temp_f(0.7), lty=2) +
  geom_segment(x = 0.7, y = temp_f(0.7), xend=0.8, yend=temp_f(0.7), lty=2) +
  geom_segment(x = 0.8, y = 0, xend = 0.8, yend = temp_f(0.8), lty=2) +
  geom_segment(x = 0.8, y = temp_f(0.8), xend=0.9, yend=temp_f(0.8), lty=2) +
  geom_segment(x = 0.9, y = 0, xend = 0.9, yend = temp_f(0.9), lty=2) +
  theme_classic() +
  labs(y = "f(x)") +
  theme(axis.ticks.x=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y=element_blank())
```


## Numerical integration: Riemann sum

Want to approximate $\int \limits_a^b f(x) dx$

\bigskip

1. Divide $[a,b]$ into $n$ subintervals of equal width $h = \dfrac{b-a}{n}$

\bigskip

2. Riemann sum approximation:

$$\int \limits_a^b f(x) dx \ \approx \ h \sum \limits_{i=0}^{n-1} f(a + ih)$$

\vspace{2cm}


## Numerical integration: trapezoid rule

Instead of a constant value in each interval, use a line:

```{r fig.width=3.5, fig.height=3, echo=F}
data.frame(x = x, y = y) |>
  ggplot() +
  geom_line(aes(x = x, y = y)) +
  geom_segment(x = 0.1, y = 0, xend = 0.1, yend = temp_f(0.1), lty=2) +
  geom_segment(x = 0.366667, y = 0, xend = 0.366667, yend = temp_f(0.366667), lty=2) +
  geom_segment(x = 0.633333, y = 0, xend = 0.633333, yend = temp_f(0.633333), lty=2) +
  geom_segment(x = 0.9, y = 0, xend = 0.9, yend = temp_f(0.9), lty=2) +
  geom_segment(x = 0.1, y = temp_f(0.1), xend=0.366667, yend=temp_f(0.366667), lty=2) +
  geom_segment(x = 0.366667, y = temp_f(0.366667), xend=0.633333, yend=temp_f(0.633333), lty=2) +
  geom_segment(x = 0.633333, y = temp_f(0.633333), xend=0.9, yend=temp_f(0.9), lty=2) +
  theme_classic() +
  labs(y = "f(x)") +
  theme(axis.ticks.x=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y=element_blank())
```


## Numerical integration: trapezoid rule

Instead of a constant value in each interval, use a line:

```{r fig.width=3.5, fig.height=3, echo=F}
data.frame(x = x, y = y) |>
  ggplot() +
  geom_line(aes(x = x, y = y)) +
  geom_segment(x = 0.1, y = 0, xend = 0.1, yend = temp_f(0.1), lty=2) +
  geom_segment(x = 0.1, y = temp_f(0.1), xend=0.2, yend=temp_f(0.2), lty=2) +
  geom_segment(x = 0.2, y = 0, xend = 0.2, yend = temp_f(0.2), lty=2) +
  geom_segment(x = 0.2, y = temp_f(0.2), xend=0.3, yend=temp_f(0.3), lty=2) +
  geom_segment(x = 0.3, y = 0, xend = 0.3, yend = temp_f(0.3), lty=2) +
  geom_segment(x = 0.3, y = temp_f(0.3), xend=0.4, yend=temp_f(0.4), lty=2) +
  geom_segment(x = 0.4, y = 0, xend = 0.4, yend = temp_f(0.4), lty=2) +
  geom_segment(x = 0.4, y = temp_f(0.4), xend=0.5, yend=temp_f(0.5), lty=2) +
  geom_segment(x = 0.5, y = 0, xend = 0.5, yend = temp_f(0.5), lty=2) +
  geom_segment(x = 0.5, y = temp_f(0.5), xend=0.6, yend=temp_f(0.6), lty=2) +
  geom_segment(x = 0.6, y = 0, xend = 0.6, yend = temp_f(0.6), lty=2) +
  geom_segment(x = 0.6, y = temp_f(0.6), xend=0.7, yend=temp_f(0.7), lty=2) +
  geom_segment(x = 0.7, y = 0, xend = 0.7, yend = temp_f(0.7), lty=2) +
  geom_segment(x = 0.7, y = temp_f(0.7), xend=0.8, yend=temp_f(0.8), lty=2) +
  geom_segment(x = 0.8, y = 0, xend = 0.8, yend = temp_f(0.8), lty=2) +
  geom_segment(x = 0.8, y = temp_f(0.8), xend=0.9, yend=temp_f(0.9), lty=2) +
  geom_segment(x = 0.9, y = 0, xend = 0.9, yend = temp_f(0.9), lty=2) +
  theme_classic() +
  labs(y = "f(x)") +
  theme(axis.ticks.x=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y=element_blank())
```


## Numerical integration: trapezoid rule


Want to approximate $\int \limits_a^b f(x) dx$

\bigskip

1. Divide $[a,b]$ into $n$ subintervals of equal width $h = \dfrac{b-a}{n}$

\bigskip

2. Trapezoid rule approximation:

$$\int \limits_a^b f(x) dx \ \approx \ \sum \limits_{i=0}^{n-1} \frac{h}{2}\left[ f(a + ih) + f(a + (i+1)h) \right]$$

$$= h \sum \limits_{i=1}^{n-1} f(a + ih) + \frac{h}{2}(f(a) + f(b))$$

\vspace{2cm}

## Your turn

Practice questions on the course website:

[https://sta379-s25.github.io/practice_questions/pq_19.html](https://sta379-s25.github.io/practice_questions/pq_19.html)

* Implement Riemann and trapezoid rules
* Start in class. You are welcome to work with others
* Practice questions are to help you practice. They are not submitted and not graded
* Solutions are posted on the course website

