---
title: "Lecture 7: Generating random variables -- acceptance-rejection sampling"
author: "Ciaran Evans"
output: beamer_presentation
---

## Previously

* Generate $U \sim Uniform(0, 1)$ with random number generator (LCG, Mersenne twister, etc)
* Some other random variables can be generated by transformations
    * Inverse transform method (if inverse cdf is easy to get)
    * Other transformations for certain variables

Today: What is another alternative?

## Example

Suppose we would like to generate $X \sim Beta(\alpha, \beta)$

$$f_X(x) = \frac{\Gamma(\alpha + \beta)}{\Gamma(\alpha)\Gamma(\beta)} x^{\alpha - 1} (1 - x)^{\beta - 1} \hspace{1cm} x \in (0, 1)$$

* Inverse transform method: $F_X(t) = \ ?$

\vspace{1cm}

* Other transformations / relationships with other distributions?

\vspace{3cm}

## Acceptance-rejection sampling: motivation

* **Want:** $X \sim Beta(\alpha, \beta)$, $f_X(x) = \frac{\Gamma(\alpha + \beta)}{\Gamma(\alpha)\Gamma(\beta)} x^{\alpha - 1} (1 - x)^{\beta - 1}$

However, it may be difficult to directly simulate from this distribution. Can you think of another distribution on $(0, 1)$ which is *easier* to simulate?

\vspace{4cm}

## Acceptance-rejection sampling: motivation

* **Want:** $X \sim Beta(\alpha, \beta)$, $f_X(x) = \frac{\Gamma(\alpha + \beta)}{\Gamma(\alpha)\Gamma(\beta)} x^{\alpha - 1} (1 - x)^{\beta - 1}$
* **Can get:** $Y \sim Uniform(0, 1)$

```{r, echo=F, message=F, warning=F, fig.width = 6, fig.height=4}
gridpts <- seq(0, 1, 0.01)
x <- dbeta(gridpts, 2.7, 6.3)
y <- dunif(gridpts)
c = 2.68
plot(gridpts,x,xlim=c(0,1),ylim=c(0,2.8),type="l",col="blue",main="",xlab="",ylab="",
     lwd=1.5)
lines(gridpts,y,col="red", lwd=1.5)
legend(.6,2,c("Target Density","Easier Density"),lty=c(1,1),col=c("blue","red"))
```

Suppose we sample $Y \sim Uniform(0, 1)$ and observe $y = 0.9$. Is it likely we would observe that draw from the Beta distribution shown here?

## Acceptance-rejection sampling: motivation

* **Want:** $X \sim Beta(\alpha, \beta)$, $f_X(x) = \frac{\Gamma(\alpha + \beta)}{\Gamma(\alpha)\Gamma(\beta)} x^{\alpha - 1} (1 - x)^{\beta - 1}$
* **Can get:** $Y \sim Uniform(0, 1)$

```{r, echo=F, message=F, warning=F, fig.width = 6, fig.height=4}
gridpts <- seq(0, 1, 0.01)
x <- dbeta(gridpts, 2.7, 6.3)
y <- dunif(gridpts)
c = 2.68
plot(gridpts,x,xlim=c(0,1),ylim=c(0,2.8),type="l",col="blue",main="",xlab="",ylab="",
     lwd=1.5)
lines(gridpts,y,col="red", lwd=1.5)
legend(.6,2,c("Target Density","Easier Density"),lty=c(1,1),col=c("blue","red"))
```

Suppose we sample $Y \sim Uniform(0, 1)$ and observe $y = 0.25$. Is it likely we would observe that draw from the Beta distribution shown here?

## Acceptance-rejection sampling

Suppose we would like to generate a continuous random variable $X$ with pdf $f$.

\vspace{6cm}

## Illustration

Target density: $f(t) = \frac{\Gamma(\alpha + \beta)}{\Gamma(\alpha)\Gamma(\beta)} x^{\alpha - 1} (1 - x)^{\beta - 1}$, $t \in (0, 1)$

Candidate density: $g(t) = 1$, $t \in (0, 1)$

```{r, echo=F, message=F, warning=F, fig.width = 6, fig.height=4}
gridpts <- seq(0, 1, 0.01)
x <- dbeta(gridpts, 2.7, 6.3)
y <- dunif(gridpts)
c = 2.68
plot(gridpts,x,xlim=c(0,1),ylim=c(0,2.8),type="l",col="blue",main="",xlab="",ylab="",
     lwd=1.5)
lines(gridpts,c*y,col="red", lwd=1.5)
#legend(.6,2,c("Target density, f(t)","c * g(t)"),lty=c(1,1),col=c("blue","red"))
```

## Illustration

Now sample $Y \sim g$ and $U \sim Uniform(0, 1)$. *Accept* $Y$ if $U \leq \frac{f(Y)}{cg(Y)}$

:::: {.columns}

::: {.column width="60%"}

```{r, echo=F, message=F, warning=F, fig.width = 6, fig.height=4, fig.align='left'}
gridpts <- seq(0, 1, 0.01)
x <- dbeta(gridpts, 2.7, 6.3)
y <- dunif(gridpts)
c = 2.68
plot(gridpts,x,xlim=c(0,1),ylim=c(0,2.8),type="l",col="blue",main="",xlab="",ylab="",
     lwd=1.5)
lines(gridpts,c*y,col="red", lwd=1.5)
abline(v=0.25)
#legend(.6,2,c("Target density, f(t)","c * g(t)"),lty=c(1,1),col=c("blue","red"))
```

:::

::: {.column width="40%"}

:::

::::

Suppose we observe $Y = 0.25$. Are we likely to *accept* or *reject* 0.25 as a sample from $f$?

## Illustration

Now sample $Y \sim g$ and $U \sim Uniform(0, 1)$. *Accept* $Y$ if $U \leq \frac{f(Y)}{cg(Y)}$

:::: {.columns}

::: {.column width="60%"}

```{r, echo=F, message=F, warning=F, fig.width = 6, fig.height=4, fig.align='left'}
gridpts <- seq(0, 1, 0.01)
x <- dbeta(gridpts, 2.7, 6.3)
y <- dunif(gridpts)
c = 2.68
plot(gridpts,x,xlim=c(0,1),ylim=c(0,2.8),type="l",col="blue",main="",xlab="",ylab="",
     lwd=1.5)
lines(gridpts,c*y,col="red", lwd=1.5)
abline(v=0.9)
#legend(.6,2,c("Target density, f(t)","c * g(t)"),lty=c(1,1),col=c("blue","red"))
```

:::

::: {.column width="40%"}

:::

::::

Suppose we observe $Y = 0.9$. Are we likely to *accept* or *reject* 0.9 as a sample from $f$?


## Illustration

Now sample $Y \sim g$ and $U \sim Uniform(0, 1)$. *Accept* $Y$ if $U \leq \frac{f(Y)}{cg(Y)}$

:::: {.columns}

::: {.column width="60%"}

```{r, echo=F, message=F, warning=F, fig.width = 6, fig.height=4, fig.align='left'}
gridpts <- seq(0, 1, 0.01)
x <- dbeta(gridpts, 2.7, 6.3)
y <- dunif(gridpts)
c = 2.68
plot(gridpts,x,xlim=c(0,1),ylim=c(0,2.8),type="l",col="blue",main="",xlab="",ylab="",
     lwd=1.5)
lines(gridpts,c*y,col="red", lwd=1.5)
abline(v=0.5)
#legend(.6,2,c("Target density, f(t)","c * g(t)"),lty=c(1,1),col=c("blue","red"))
```

:::

::: {.column width="40%"}

:::

::::

Suppose we observe $Y = 0.5$. Are we likely to *accept* or *reject* 0.5 as a sample from $f$?



## Illustration

* $Y \sim g$ and $U \sim Uniform(0, 1)$
* *Accept* $Y$ if $U \leq \frac{f(Y)}{cg(Y)}$

:::: {.columns}

::: {.column width="70%"}

```{r, echo=F, message=F, warning=F, fig.width = 7, fig.height=5, fig.align='left'}
gridpts <- seq(0, 1, 0.01)
x <- dbeta(gridpts, 2.7, 6.3)
y <- dunif(gridpts)
c = 2.68
plot(gridpts,x,xlim=c(0,1),ylim=c(0,2.8),type="l",col="blue",main="",xlab="",ylab="",
     lwd=1.5)
lines(gridpts,c*y,col="red", lwd=1.5)
#legend(.6,2,c("Target density, f(t)","c * g(t)"),lty=c(1,1),col=c("blue","red"))
```

:::

::: {.column width="30%"}

:::

::::


## Your turn

Practice questions on the course website:

[https://sta379-s25.github.io/practice_questions/pq_7.html](https://sta379-s25.github.io/practice_questions/pq_7.html)

* Implement acceptance-rejection sampling for the beta example
* Start in class. You are welcome to work with others
* Practice questions are to help you practice. They are not submitted and not graded
* Solutions are posted on the course website

**Next time:** formal proof that this sampling procedure works!


