---
title: "Lecture 23: Changing the range of integration"
author: "Ciaran Evans"
output: beamer_presentation
editor_options: 
  chunk_output_type: console
---

## Gauss-Legendre quadrature

* Let $x_1,...,x_n \in (-1, 1)$ be the $n$ roots of the $n$th Legendre polynomial $p_n$

\bigskip

* Use $x_1,...,x_n$ as quadrature nodes to approximate integrals:
$$\int \limits_{-1}^1 f(x)dx \approx  \sum \limits_{i=1}^n w_i f(x_i) \hspace{1cm} w_i = \int\limits_{-1}^1 L_{n,i}(x) dx$$

\bigskip

* If $f(x)$ is a polynomial of degree $\leq 2n-1$, approximation is **exact**:
$$\int \limits_{-1}^1 f(x)dx =  \sum \limits_{i=1}^n w_i f(x_i)$$


## Changing the range of integration

Gauss-Legendre quadrature allows us to approximate $\int \limits_{-1}^1 f(x)dx$.

\bigskip

**Question:** What should I do if I want to approximate
$$\int \limits_a^b f(x) dx$$
for a finite interval $[a,b]$?

\vspace{4cm}

## Integrating over an infinite range

Integrals in statistics often involve an infinite range. For example, standard normal cdf:
$$\int \limits_{-\infty}^t \frac{1}{\sqrt{2 \pi}} e^{-\frac{1}{2}x^2} dx$$

**Question:** How could we use Gauss-Legendre quadrature to approximate the integral?

\vspace{4cm}

## Integrating over an infinite range: truncation

The standard normal density is mostly concentrated around 0, so for many values of $t$

$$\int \limits_{-\infty}^t \frac{1}{\sqrt{2 \pi}} e^{-\frac{1}{2}x^2} dx \approx \int \limits_{-5}^t \frac{1}{\sqrt{2 \pi}} e^{-\frac{1}{2}x^2} dx$$

```{r}
pnorm(-1)
pnorm(-1) - pnorm(-5)
```

## Integrating over an infinite range: truncation

```{r}
library(rootSolve)
p4_roots <- uniroot.all(function(x){
  (1/8) * (35*x^4 - 30*x^2 + 3)}, 
                        c(-1, 1), tol=1e-12)
weights <- c((18 - sqrt(30))/36, 
             (18 + sqrt(30))/36,
             (18 + sqrt(30))/36,
             (18 - sqrt(30))/36)

a <- -5; b <- -1
(b - a)/2*sum(weights*dnorm((b-a)/2*p4_roots + (a+b)/2))

pnorm(-1)
```

## Integrating over an infinite range: transformation

Find a transformation $x = h(u)$ such that

$$\int \limits_a^b f(x)dx = \int \limits_{-1}^1 f(h(u)) h'(u)du$$

**Question:** For $$\int \limits_{-\infty}^t \frac{1}{\sqrt{2 \pi}} e^{-\frac{1}{2}x^2} dx$$ what transformations could I consider?

## Integrating over an infinite range: transformation

$$\int \limits_{-\infty}^t \frac{1}{\sqrt{2 \pi}} e^{-\frac{1}{2}x^2} dx$$

Let $x = t + \log\left( \dfrac{u+1}{2} \right)$

\vspace{4cm}

## Integrating over an infinite range: transformation

```{r}
f <- function(u){
  exp(-0.5*(-1 + log(0.5*(u+1)))^2)/(u+1)
}

sum(weights * f(p4_roots))/sqrt(2*pi)

pnorm(-1)
```

## Gauss-Hermite quadrature

Lots of integrals in statistics involve the normal distribution, and so look like

$$\int \limits_{-\infty}^\infty f(x) e^{-\frac{1}{2}x^2} dx$$
**Gauss-Hermite quadrature** is a quadrature rule that is good at these types of integrals:

$$\int \limits_{-\infty}^\infty f(x) e^{-\frac{1}{2}x^2} dx \approx \sum \limits_{i=1}^n w_i f(x_i)$$

* Need to choose the $x_i$ and $w_i$ differently to Gauss-Legendre quadrature

## Gauss-Hermite quadrature

Gauss-Legendre quadrature: $\int \limits_{-1}^1 f(x) dx \approx \sum \limits_{i=1}^n w_i f(x_i)$

* $w_i = \int \limits_{-1}^1 L_{n,i}(x)dx$
* $x_1,...,x_n$ are the roots of the $n$th Legendre polynomial $p_n$. Legendre polynomials satisfy $\int \limits_{-1}^1 (c_0 + c_1x + \cdots + c_{n-1}x^{n-1}) p_n(x) dx$


Gauss-Hermite quadrature: $\int \limits_{-\infty}^\infty f(x) e^{-\frac{1}{2}x^2} dx \approx \sum \limits_{i=1}^n w_i f(x_i)$

* $w_i =$


## Gauss-Hermite quadrature

Gauss-Legendre quadrature: $\int \limits_{-1}^1 f(x) dx \approx \sum \limits_{i=1}^n w_i f(x_i)$

* $w_i = \int \limits_{-1}^1 L_{n,i}(x)dx$
* $x_1,...,x_n$ are the roots of the $n$th Legendre polynomial $p_n$. Legendre polynomials satisfy $\int \limits_{-1}^1 (c_0 + c_1x + \cdots + c_{n-1}x^{n-1}) p_n(x) dx$


Gauss-Hermite quadrature: $\int \limits_{-\infty}^\infty f(x) e^{-\frac{1}{2}x^2} dx \approx \sum \limits_{i=1}^n w_i f(x_i)$

* $w_i = \int \limits_{-\infty}^\infty L_{n,i}(x) e^{-\frac{1}{2}x^2} dx$
* $x_1,...,x_n$ are the roots of the $n$th **Hermite** polynomial $h_n$. Hermite polynomials satisfy $\int \limits_{-\infty}^\infty (c_0 + c_1x + \cdots + c_{n-1}x^{n-1}) h_n(x) e^{-\frac{1}{2}x^2} dx = 0$

## Example

Hermite polynomial for $n=2$: $h_2(x) = x^2 - 1$

* Roots of $h_2$:

\bigskip

* Weights: $w_i = \int \limits_{-\infty}^\infty L_{n,i}(x) e^{-\frac{1}{2}x^2} dx$

\vspace{5cm}


## Example

$$\int \limits_{-\infty}^\infty x^2 e^{-\frac{1}{2}x^2} dx = \hspace{3cm}$$

Gauss-Hermite quadrature with $n=2$: $w_1f(x_1) + w_2f(x_2) =$

\vspace{4cm}

## Example

$$\int \limits_{-\infty}^\infty x^2 e^{-\frac{1}{2}x^2} dx = \sqrt{2 \pi}$$

Gauss-Hermite quadrature with $n=2$: 
$$w_1f(x_1) + w_2f(x_2) = \sqrt{\frac{\pi}{2}} \cdot  (-1)^2 + \sqrt{\frac{\pi}{2}} \cdot  (1)^2 = \sqrt{2 \pi}$$

```{r}
nodes <- c(-1, 1)
weights <- c(sqrt(pi/2), sqrt(pi/2))
sum(weights * nodes^2)
sqrt(2*pi)
```

## Your turn

Try Gauss-Hermite quadrature for calculating expectations of functions of normal distributions:

[https://sta379-s25.github.io/practice_questions/pq_23.html](https://sta379-s25.github.io/practice_questions/pq_23.html)

* Start in class
* You are welcome and encouraged to work with your neighbors
* Solutions posted on course website


