---
title: "Lecture 11: Nelder-Mead method"
author: "Ciaran Evans"
output: beamer_presentation
editor_options: 
  chunk_output_type: console
---

## Recap: optimization

**Definition:** *Optimization* is the problem of finding values that minimize or maximize some function.

**Example:** 

$$RSS(\beta_0, \beta_1) = \sum \limits_{i=1}^n (\text{Weight}_i - \beta_0 - \beta_1 \text{WingLength}_i)^2$$

* $RSS(\beta_0, \beta_1)$ is a function of $\beta_0$ and $\beta_1$
* We want to find the values of $\beta_0$ and $\beta_1$ that *minimize* this function

## Last time: Compass search overview (in 2 dimensions)

To minimize some function $f(\beta_0, \beta_1)$:

1. Choose an initial guess $(\beta_0^{(0)}, \beta_1^{(0)})$ and initial step size $\Delta_0$
2. Evaluate $f$ at the points
    * $(\beta_0^{(0)}, \beta_1^{(0)})$
    * $(\beta_0^{(0)}, \beta_1^{(0)} \pm \Delta_0)$
    * $(\beta_0^{(0)} \pm \Delta_0, \beta_1^{(0)})$
3. If $f$ is smaller at one of the new points: move to the smallest value, update to $(\beta_0^{(1)}, \beta_1^{(1)})$
4. Otherwise: $\Delta_{k+1} = 0.5 \Delta_k$ (shrink step size and try again)
5. Repeat

\vspace{2cm}

## Downsides of compass search

Downsides of the compass search algorithm given on the previous slide:

* Can require many steps
* Only considers specific search directions
* Step size only shrinks; if we find a promising direction, can't take bigger steps
* Choosing a direction requires many evaluations of $f$
    * In 2 dimensions (e.g. $\beta_0$ and $\beta_1$), requires 4 evaluations of $f$
    * In $d$ dimensions ($\beta_0, \beta_1,...,\beta_{d-1}$), requires 2$d$ evaluations of $f$
    
**Question:** How could you modify the algorithm to address some of the issues here?

## A modified compass search algorithm


## Alternative: Nelder-Mead method

In R, standard function for performing optimization is `optim`:

```{r, eval=F}
?optim
```

```
optim(par, fn, gr = NULL, ...,
      method = c("Nelder-Mead", "BFGS", "CG", "L-BFGS-B", "SANN",
                 "Brent"),
      lower = -Inf, upper = Inf,
      control = list(), hessian = FALSE)
```


* Another derivative-free optimization method
* **Very** widely used
    * Original 1965 paper has 40,000+ citations
    * Can find many examples of use in biology, medicine, physics, engingeering, etc.


## Nelder-Mead method (in 2 dimensions)

Start with 3 initial points, evaluate function $f$ at each point:

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=4, fig.height=2.5}
library(Stat2Data)
library(ggrepel)
data("Sparrows")

rss <- function(beta){
  sum((Sparrows$Weight - beta[1] - beta[2]*Sparrows$WingLength)^2)
}

x1 <- c(0.3598633, 0.5937500, 0.4346924)
x2 <- c(0.5498047, 0.5625000, 0.4870605)

cur_simplex <- cbind(x1, x2)

rss_vals <- apply(cbind(x1, x2), 1, rss)

set.seed(3)
data.frame(x1, x2, rss_vals) |>
  ggplot(aes(x = x1, y = x2, label=round(rss_vals))) +
  geom_point() +
  #geom_point(aes(x = 0.5, y = 0.5))+
  geom_polygon(color="black", fill=NA) +
  theme_classic() +
  geom_text_repel(size=3, force_pull = 0.5) +
  scale_x_continuous(limits=c(0.1, 0.75)) +
  scale_y_continuous(limits = c(0.4, 0.6)) +
  labs(x = "Beta0", y = "Beta1")
```

**Question:** Where should I search next?


## Nelder-Mead method (in 2 dimensions)

First, calculate centroid of the vertices (except the worst one)

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=4, fig.height=2.5}
x1 <- c(0.3598633, 0.5937500, 0.4346924)
x2 <- c(0.5498047, 0.5625000, 0.4870605)

cur_simplex <- cbind(x1, x2)

rss_vals <- apply(cbind(x1, x2), 1, rss)
cur_simplex <- cur_simplex[order(rss_vals),]
centroid <- colMeans(cur_simplex[1:2,])

set.seed(3)
data.frame(x1, x2, rss_vals) |>
  ggplot(aes(x = x1, y = x2, label=round(rss_vals))) +
  geom_point() +
  geom_point(aes(x = centroid[1], y = centroid[2])) +
  geom_polygon(color="black", fill=NA) +
  theme_classic() +
  geom_text_repel(size=3, force_pull = 0.5) +
  scale_x_continuous(limits=c(0.1, 0.75)) +
  scale_y_continuous(limits = c(0.4, 0.6)) +
  labs(x = "Beta0", y = "Beta1")
```


## Reflection

Now reflect the worst vertex over the centroid:

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=4, fig.height=2.5}
x1 <- c(0.3598633, 0.5937500, 0.4346924)
x2 <- c(0.5498047, 0.5625000, 0.4870605)

cur_simplex <- cbind(x1, x2)

rss_vals <- apply(cbind(x1, x2), 1, rss)
cur_simplex <- cur_simplex[order(rss_vals),]
centroid <- colMeans(cur_simplex[1:2,])
reflection = centroid + (centroid - cur_simplex[3,])

set.seed(3)
data.frame(x1, x2, rss_vals) |>
  ggplot(aes(x = x1, y = x2, label=round(rss_vals))) +
  geom_point() +
  geom_point(aes(x = centroid[1], y = centroid[2])) +
  geom_point(aes(x = reflection[1], y = reflection[2])) +
  geom_polygon(color="black", fill=NA) +
  theme_classic() +
  geom_text_repel(size=3, force_pull = 0.5) +
  scale_x_continuous(limits=c(0.1, 0.75)) +
  scale_y_continuous(limits = c(0.4, 0.6)) +
  labs(x = "Beta0", y = "Beta1") +
  geom_segment(aes(x = reflection[1], y = reflection[2], 
                   xend= cur_simplex[3,1], yend = cur_simplex[3,2]), lty=2)
```




## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

Initial points:

```{r, echo=F, fig.align='center', fig.width=4, fig.height=2.5}
plot_triangle <- function(cur_simplex, xlim=NA, ylim=NA,
                          quiet = F){
  
  if(any(is.na(xlim))){
    buffer <- diff(range(cur_simplex[,1]))
    xlim = c(min(cur_simplex[,1]) - 0.5*buffer, 
             max(cur_simplex[,1]) + 0.5*buffer)
  }
  if(any(is.na(ylim))){
    buffer <- diff(range(cur_simplex[,2]))
    ylim = c(min(cur_simplex[,2]) - 0.5*buffer, 
             max(cur_simplex[,2]) + 0.5*buffer)
  }
  if(quiet){
    data.frame(x1 = cur_simplex[,1], 
               x2 = cur_simplex[,2]) |>
      ggplot(aes(x = x1, y = x2)) +
      geom_point() +
      geom_polygon(color="black", fill=NA) +
      theme_classic() +
      scale_x_continuous(limits=xlim) +
      scale_y_continuous(limits = ylim) +
      labs(x = "Beta0", y = "Beta1")
  } else {
    rss_vals <- apply(cur_simplex, 1, rss)
    data.frame(x1 = cur_simplex[,1], 
               x2 = cur_simplex[,2],
               rss_vals) |>
      ggplot(aes(x = x1, y = x2, label=rss_vals)) +
      geom_point() +
      geom_polygon(color="black", fill=NA) +
      theme_classic() +
      geom_text_repel(size=3, force_pull = 0.5) +
      scale_x_continuous(limits=xlim) +
      scale_y_continuous(limits = ylim) +
      labs(x = "Beta0", y = "Beta1")
  }
  
}


nm_step <- function(cur_simplex){
  rss_vals <- apply(cur_simplex, 1, rss)
  cur_simplex <- cur_simplex[order(rss_vals),]
  rss_vals <- sort(rss_vals)
  centroid <- colMeans(cur_simplex[1:2,])
  reflection = centroid + (centroid - cur_simplex[3,])
  
  if(rss(reflection) < rss_vals[2] && rss(reflection) >= rss_vals[1]){
    # reflection
    cur_simplex <- rbind(cur_simplex[1:2,], reflection)
    print("reflection")
  } else if(rss(reflection) < rss_vals[1]){
    # expansion?
    expansion = centroid + 2*(centroid - cur_simplex[3,])
    if(rss(expansion) < rss(reflection)){
      cur_simplex <- rbind(cur_simplex[1:2,], expansion)
      print("expansion")
    } else {
      cur_simplex <- rbind(cur_simplex[1:2,], reflection)
      print("reflection")
    }
  } else if(rss(reflection) < rss_vals[3]){
    # contract outside
    co <- centroid + 0.5*(reflection - centroid)
    if(rss(co) < rss(reflection)){
      cur_simplex <- rbind(cur_simplex[1:2,], co)
      print("contract outside")
    } else {
      # shrink
      cur_simplex[2,] <- cur_simplex[1,] + 0.5*(cur_simplex[2,] - cur_simplex[1,])
      cur_simplex[3,] <- cur_simplex[1,] + 0.5*(cur_simplex[3,] - cur_simplex[1,])
      print("shrink")
    }
  } else {
    # contract inside
    ci <- centroid + 0.5*(cur_simplex[3,] - centroid)
    if(rss(ci) < rss_vals[3]){
      cur_simplex <- rbind(cur_simplex[1:2,], ci)
      print("contract inside")
    } else {
      # shrink
      cur_simplex[2,] <- cur_simplex[1,] + 0.5*(cur_simplex[2,] - cur_simplex[1,])
      cur_simplex[3,] <- cur_simplex[1,] + 0.5*(cur_simplex[3,] - cur_simplex[1,])
      print("shrink")
    }
  }
  
  return(cur_simplex)
}




x1 <- c(0, 0.5, 1)
x2 <- c(0, 1, 0)

cur_simplex <- cbind(x1, x2)

plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5))
```


## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5))
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5))
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5))
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5))
```


## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead in action

$\widehat{\text{Weight}}_i = 1.3655 + 0.4674 \text{WingLength}_i$

```{r echo=F, fig.align='center', fig.width=4, fig.height=2.5}
cur_simplex <- nm_step(cur_simplex)
plot_triangle(cur_simplex, xlim=c(-0.5, 1.5), ylim = c(-0.5, 1.5), quiet=T)
```

## Nelder-Mead summary

* Explores *away* from the current worst point
    * *Reflection* tries a point in the new direction
* *Expansion* allows us to increase the "step size" if the new direction is particularly promising
* *Contraction* and *shrinking* allows us to decrease the "step size" if we are moving too far




