---
title: "Lecture 3: Functions"
author: "Ciaran Evans"
output: beamer_presentation
---

## Last time

```{r, eval=F}
n <- 1000
m = 2^32
a = 1664525
c = 1013904223

x0 <- 1

x <- rep(NA, n)
x[1] <- x0
for(i in 2:n){
  x[i] <- (a * x[i-1] + c) %% m
}

u <- x/m
```


What if I wanted to run the LCG with a different seed, or a different number of samples, or different parameters?

## LCG code changing the settings

```{r, eval=F}
n <- 1000
m = 2^32
a = 1664525
c = 1013904223

x0 <- 1

...

n <- 5000
m = 1024
a = 1
c = 1

x0 <- 12

...

```

## Functions

```{r}
my_lcg <- function(n, x0, m, a, c){
  x <- rep(NA, n)
  x[1] <- x0
  for(i in 2:n){
    x[i] <- (a * x[i-1] + c) %% m
  }
  return(x/m)
}

head(my_lcg(n = 1000, x0 = 1, m = 2^32, 
            a = 1664525, c = 1013904223))

head(my_lcg(n = 5000, x0 = 12, m = 1024, a = 1, c = 1))
```

## Function defaults

```{r}
my_lcg <- function(n, x0, m = 2^32, 
                   a = 1664525, c = 1013904223){
  x <- rep(NA, n)
  x[1] <- x0
  for(i in 2:n){
    x[i] <- (a * x[i-1] + c) %% m
  }
  return(x/m)
}

my_lcg(n = 5, x0 = 1, m = 2^32, 
       a = 1664525, c = 1013904223)

my_lcg(n = 5, x0 = 1)
```

## What if we don't want to specify a seed?

* In R, functions for simulating random variables don't *require* us to specify a seed:

```{r}
runif(5)
```

* But even if we don't manually set a seed, R's random number generators still require one!

How might R choose a seed, if we don't specify one ourselves?


## Getting the system time

```{r}
Sys.time()
```


## Getting the system time

```{r}
Sys.time() |>
  lubridate::second()
```

```{r}
(Sys.time() |> lubridate::second()) * 100000
```


## What if we don't want to specify a seed?

```{r}
my_lcg <- function(n, x0 = NA, m = 2^32, 
                   a = 1664525, c = 1013904223){
  if(is.na(x0)){
    x0 <- ((Sys.time() |> 
              lubridate::second()) * 100000) %% m
  }
  
  x <- rep(NA, n)
  x[1] <- x0
  for(i in 2:n){
    x[i] <- (a * x[i-1] + c) %% m
  }
  return(x/m)
}

my_lcg(n = 5)
```

## What if we don't want to specify a seed?

```{r}
my_lcg(n = 5)
```

```{r, include=F}
Sys.sleep(3)
```


```{r}
my_lcg(n = 5)
```



## Homework 1

[https://sta379-s25.github.io/homework/hw1.html](https://sta379-s25.github.io/homework/hw1.html)

* Function practice and probability review
* Accept and submit coding portion of assignment on GitHub Classroom
* Collaboration encouraged on homework, but everyone must submit their own work and acknowledge collaborators


