---
title: "Lecture 4: Some limitations of R, and intro to C++"
author: "Ciaran Evans"
output: beamer_presentation
---

## Previously: Linear congruential generator

$$x_{n+1} = (ax_n + c) \ \text{mod} \ m$$

While a classic method, R does not use an LCG to generate its random numbers:

```{r, eval=F}
?.Random.seed
```

\begin{verbatim}
Details

The currently available RNG kinds are given below...
The default is "Mersenne-Twister".
\end{verbatim}

## What is the Mersenne Twister?

We won't go through all the details right now, but here is the core of the algorithm: we generate a sequence of integers by

$$x_{k} = x_{k - (n-m)} \oplus ((x_{k-n}^u | x_{k-(n-1)}^l) A)$$

$$x A = \begin{cases} x \gg 1 & x_{[0]} = 0 \ \ (x \text{ is even}) \\
(x \gg 1) \oplus a & x_{[0]} = 1 \ \ (x \text{ is odd}) \end{cases}$$

Here, $\oplus$, $\gg$, and $|$ all represent **bitwise** operations on $x$

## Example: bitwise shift

Consider a 4-bit integer:

$$3 = 0 0 1 1$$

$$3 \gg 1 = 0001 = 1$$

$$3 \gg 2 = 0000= 0$$

$$3 \ll 1 = 0 1 1 0 = 6$$


## An example in R

$$3 \gg 1 = 1$$

```{r}
bitwShiftR(3, 1)
```

$$3 \ll 1 = 6$$

```{r}
bitwShiftL(3, 1)
```


## Another example in R

```{r}
bitwShiftL(2147483647, 1)
```



Wait... a left shift of 1 should *double* our number, right? How did we get $-2$??


## Integers in R

Integers in R are *signed*, *32-bit* integers:

$$0 = 00000000000000000000000000000000$$

$$1 = 00000000000000000000000000000001$$

$$2147483647 = 2^{31} - 1 = 01111111111111111111111111111111$$

```{r}
.Machine$integer.max
```

## 32-bit signed integers

$$2147483647 = 2^{31} - 1 = 01111111111111111111111111111111$$

$$? = 10000000000000000000000000000000$$


## 32-bit signed integers

$$2147483647 = 2^{31} - 1 = 01111111111111111111111111111111$$

$$-2147483648 = 10000000000000000000000000000000$$

$$-2147483647 = 10000000000000000000000000000001$$

$$-2147483646 = 10000000000000000000000000000010$$

etc.

$$-2 = 11111111111111111111111111111110$$

$$-1 = 11111111111111111111111111111111$$

## Integers in R

$$2147483647 = 01111111111111111111111111111111$$

$$2147483647 \ll 1 = 11111111111111111111111111111110 = -2$$

## What behavior do we *want*?

Random number generators produce numbers between 0 and 1 $\Longrightarrow$ we don't want any negative numbers. Instead, we want:

$$2147483647 = 01111111111111111111111111111111$$

$$2147483648 = 10000000000000000000000000000000$$

...

$$4294967295 = 11111111111111111111111111111111$$

That is, we want 32-bit **unsigned** integers (values between 0 and $2^{32} -1$)

## Types in R

6 basic scalar types in R:

* logical (`TRUE` or `FALSE`)
* double (for decimal numbers)
* integer (32-bit, signed)
* character (i.e., strings)
* complex (for complex numbers)
* raw (specifically for working with binary data)

**Problem:** No option for *unsigned* integers!

## LCG in C++

Here is an implementation of the LCG in C++

```{Rcpp, eval=F}
arma::vec my_lcgC(int n, uint32_t x0, 
                  uint64_t m = 4294967296, 
                  uint32_t a = 1664525, 
                  uint32_t c = 1013904223){
  arma::uvec x(n);

  x[0] = x0;
  for(int i = 1; i < n; i++){
    x[i] = (a*x[i-1] + c) % m;
  }
  
  arma::vec u = arma::conv_to<arma::vec>::from(x);
  return u/m;
}
```

How does this compare to R code?

## Some data types in C++

* `int`: signed 32-bit integers
* `uint32_t`: unsigned 32-bit integers
* `uint64_t`: unsigned 64-bit integers
* `bool`: boolean (`true` or `false`)
* `double`: double-precision floating point number (for decimals)
* `arma::vec`: vector (of doubles) from Armadillo library
* `arma::uvec`: vector (of unsigned integers) from Armadillo library

## Another example

```{Rcpp, eval=F}
double sumC(arma::vec x) {
  int n = x.n_elem;
  double total = 0;
  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total;
}
```

What is this code doing?

## Comparing R and C++ speed

\small

```{r}
Rcpp::cppFunction('double sumC(arma::vec x) {
  int n = x.n_elem;
  double total = 0;
  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total;
}', depends = "RcppArmadillo")

x <- rnorm(10000)
bench::mark(
  sum(x),
  sumC(x)
)
```
## Comparing R and C++ speed

```{r, include=F}
Rcpp::cppFunction('arma::vec my_lcgC(int n, uint32_t x0, 
                  uint64_t m = 4294967296, 
                  uint32_t a = 1664525, 
                  uint32_t c = 1013904223){
  arma::uvec x(n);

  x[0] = x0;
  for(int i = 1; i < n; i++){
    x[i] = (a*x[i-1] + c) % m;
  }
  
  arma::vec u = arma::conv_to<arma::vec>::from(x);
  return u/m;
}', depends = "RcppArmadillo")

my_lcg <- function(n, x0, m = 2^32, 
                   a = 1664525, c = 1013904223){
  x <- rep(NA, n)
  x[1] <- x0
  for(i in 2:n){
    x[i] <- (a * x[i-1] + c) %% m
  }
  return(x/m)
}
```

```{r}
bench::mark(
  my_lcg(1000, 1),
  my_lcgC(1000, 1),
  check=F
)
```

## Some key points

* C++ can be faster than an equivalent implementation in R, especially loops/iteration
* C++ can be more general-purpose, and provides a wider variety of certain data types
* C++ *always* needs to know the type of an object
    * This is true for inputs, outputs, *and* any variables you create
* In C++, indexing begins at 0
* C++ needs a `;` at the end of each line
* The Armadillo library provides many useful objects and functions that behave similarly to R counterparts

## Example: Correlation

Suppose we have a sample $(X_1, Y_1),...,(X_n, Y_n)$ of $n$ observations collected on two variables, $X$ and $Y$. The *sample correlation* is given by

$$\dfrac{ \sum \limits_{i=1}^n (X_i - \bar{X})(Y_i - \bar{Y})}{\left( \sum \limits_{i=1}^n (X_i - \bar{X})^2 \right)^{1/2} \left( \sum \limits_{i=1}^n (Y_i - \bar{Y})^2 \right)^{1/2}}$$

Suppose I want to write a function to calculate the sample correlation in C++.

* What should the inputs be?
* What should the output be?
* What steps do I need to do inside the function?

## Example: Correlation

Beginning to define the function:

```{Rcpp, eval=F}
double cor_C(arma::vec x, arma::vec y) {
  



}
```

## Example: Correlation

```{Rcpp, eval=F}
double cor_C(arma::vec x, arma::vec y) {
  arma::vec diffsx = x - mean(x);
  arma::vec diffsy = y - mean(y);
  return sum(diffsx % diffsy)/
    (sqrt(sum(square(diffsx))) * 
      sqrt(sum(square(diffsy))));
}
```

## Get the function into R

```{r}
Rcpp::cppFunction('double cor_C(arma::vec x, arma::vec y){
  arma::vec diffsx = x - mean(x);
  arma::vec diffsy = y - mean(y);
  return sum(diffsx % diffsy)/
    (sqrt(sum(square(diffsx))) * 
      sqrt(sum(square(diffsy))));
}', depends = "RcppArmadillo")

x <- rnorm(100)
y <- rnorm(100)

cor_C(x, y) # our version
cor(x, y)   # existing R version
```




## Your turn

Practice questions on the course website:

[https://sta379-s25.github.io/practice_questions/pq_4.html](https://sta379-s25.github.io/practice_questions/pq_4.html)

* Practice writing short functions in C++
* Start in class. You are welcome to work with others
* Practice questions are to help you practice. They are not submitted and not graded
* Solutions are posted on the course website
