---
title: "Lecture 31: Gaussian mixture models with multivariate data"
author: "Ciaran Evans"
output: beamer_presentation
editor_options: 
  chunk_output_type: console
---

## Course logistics

* HW 8 released, due Monday 4/28

\bigskip

* This week: wrap up Gaussian mixtures and EM algorithm

\bigskip

* Course evals: Wednesday during class

\bigskip

* Monday 4/28: wrap-up work day

## Previously

```{r, fig.width=3.5, fig.height=2.5, message=F, warning=F, echo=F}
library(MASS)
library(tidyverse)
library(mixtools)

geyser |>
  ggplot(aes(x = waiting)) +
  geom_histogram(fill="transparent", color="black") +
  labs(x = "Time between eruptions (minutes)") +
  theme_classic()
```

## Previously: Gaussian mixture model

* Observe data $X_1,...,X_n$

\bigskip

* Assume each observation $i$ comes from one of $k$ groups. Let $Z_i \in \{1,...,k\}$ denote the group assignment 
    * The group $Z$ is an unobserved (**latent**) variable

**Model:** 

* $P(Z_i = j) = \lambda_j$

\bigskip

* $X_i | (Z_i = j) \sim N(\mu_j, \sigma_j^2)$

## Posterior probabilities and parameter estimation

* **If** we know the parameters $\lambda$, $\mu$, $\sigma$, we can calculate posterior probabilities:

\vspace{-0.3cm}

$$P(Z_i = j | X_i) = \dfrac{\lambda_j f(X_i | Z_i = j)}{\lambda_1 f(X_i | Z_i = 1) + \cdots + \lambda_k f(X_i |Z_i = k)}$$


* **If** we know the posterior probabilities, we can estimate the model parameters $\lambda$, $\mu$, and $\sigma$:

    * $\widehat{\lambda}_j = \frac{1}{n} \sum \limits_{i=1}^n P(Z_i = j | X_i)$
    * $\widehat{\mu}_j = \dfrac{\sum \limits_{i=1}^n X_i P(Z_i = j | X_i)}{\sum \limits_{i=1}^n {P(Z_i = j | X_i)}}$
    * $\widehat{\sigma}_j = \sqrt{\dfrac{\sum \limits_{i=1}^n (X_i - \widehat{\mu}_j)^2 P(Z_i = j | X_i)}{\sum \limits_{i=1}^n {P(Z_i = j | X_i)}}}$


## Putting everything together

**Model:** $P(Z_i = j) = \lambda_j \hspace{1cm} X_i | (Z_i = j) \sim N(\mu_j, \sigma_j^2)$

**Parameters:** $\lambda = (\lambda_1,...,\lambda_k)$, $\mu = (\mu_1,...,\mu_k)$, $\sigma = (\sigma_1,...,\sigma_k)$

**Estimation:**

1. Initialize parameter guesses $\lambda^{(0)}$, $\mu^{(0)}$, $\sigma^{(0)}$

2. Given current parameter estimates, compute $P^{(0)}(Z_i = j | X_i)$ for all $i, j$

3. Given current posterior probabilities $P^{(0)}(Z_i = j | X_i)$, update parameter estimates to $\lambda^{(1)}$, $\mu^{(1)}$, $\sigma^{(1)}$

4. Iterate: repeat steps 2--3 until convergence


## Multivariate data

Penguin data:

```{r echo=F, message=F, warning=F, fig.width=3.25, fig.height=2.5}
library(palmerpenguins)

penguins |>
  ggplot(aes(x = bill_depth_mm, y = bill_length_mm)) +
  geom_point(size=1) +
  labs(x = "Bill depth (mm)", y = "Bill length (mm)") +
  theme_classic()
```

**Question:** What do you notice about this scatterplot?

## Multivariate data


```{r echo=F, message=F, warning=F, fig.width=3.25, fig.height=3}
library(palmerpenguins)

penguins |>
  ggplot(aes(x = bill_depth_mm, y = bill_length_mm, color = species)) +
  geom_point(size=1) +
  labs(x = "Bill depth (mm)", y = "Bill length (mm)") +
  theme_classic() +
  theme(legend.position = "bottom")
```

## Multivariate data

```{r echo=F, message=F, warning=F, fig.width=3.25, fig.height=2.5}
data.frame(x = geyser$waiting[2:299], y = geyser$duration[1:298]) |>
  ggplot(aes(x = y, y = x)) +
  geom_point(size = 0.5) +
  labs(y = "Time between eruptions (minutes)",
       x = "Duration of eruption (minutes)") +
  theme_classic()
```

**Question:** How should we generalize our Gaussian mixture model to multivariate data?

## Multivariate normal distribution

**Definition:** Let $X = (X_1,...,X_k)^T$. We say that $X \sim N(\mu, \Sigma)$ if for any ${\bf a} \in \mathbb{R}^k$, ${\bf a}^T X$ follows a (univariate) normal distribution.

* $\mu =$

\vspace{2cm}

* $\Sigma =$

\vspace{2cm}

## Multivariate normal distribution

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=3.5, fig.height=3.5}
library(MASS)
library(patchwork)
library(tidyverse)
y <- mvrnorm(1000, mu=c(0, 0), Sigma = diag(2))

p1 <- data.frame(x1 = y[,1], x2 = y[,2]) |>
  ggplot(aes(x = x1, y = x2)) +
  geom_point(alpha = 0.2, size=0.3) +
  stat_ellipse(type="norm", level = 0.5) +
  stat_ellipse(type="norm", level = 0.7) +
  stat_ellipse(type="norm", level = 0.9) +
  theme_classic()

y <- mvrnorm(1000, mu=c(0, 0), Sigma = diag(c(1, 5)))

p2 <- data.frame(x1 = y[,1], x2 = y[,2]) |>
  ggplot(aes(x = x1, y = x2)) +
  geom_point(alpha = 0.2, size=0.3) +
  stat_ellipse(type="norm", level = 0.5) +
  stat_ellipse(type="norm", level = 0.7) +
  stat_ellipse(type="norm", level = 0.9) +
  theme_classic() +
  scale_x_continuous(limits = c(-6, 6)) +
  scale_y_continuous(limits = c(-6, 6))

y <- mvrnorm(1000, mu=c(0, 0), Sigma = matrix(c(1, 0.75, 0.75, 1), nrow=2))

p3 <- data.frame(x1 = y[,1], x2 = y[,2]) |>
  ggplot(aes(x = x1, y = x2)) +
  geom_point(alpha = 0.2, size=0.3) +
  stat_ellipse(type="norm", level = 0.5) +
  stat_ellipse(type="norm", level = 0.7) +
  stat_ellipse(type="norm", level = 0.9) +
  theme_classic() +
  scale_x_continuous(limits = c(-3, 3)) +
  scale_y_continuous(limits = c(-3, 3))

(p1 + p2)/(p3 + plot_spacer())

#alpha_levels <- c(0.5, 0.9)
#contour_data <- data.table::rbindlist(lapply(alpha_levels,ellipse,x=diag(2),
#      scale=c(1,1),  ## needed for positional matching
#      centre=c(0, 0)))
```

## Multivariate Gaussian mixture model


```{r, include=F}
old_faithful <- cbind(geyser$duration[1:298], geyser$waiting[2:299])
em_res <- mvnormalmixEM(old_faithful, 
                        lambda = c(0.5, 0.5), k=2)
```

```{r, eval=F}
em_res <- mvnormalmixEM(old_faithful, 
                        lambda = c(0.5, 0.5), k=2)
```


```{r echo=F, message=F, warning=F, fig.width=2.5, fig.height=1.7}
data.frame(x = geyser$waiting[2:299], y = geyser$duration[1:298]) |>
  ggplot(aes(x = y, y = x)) +
  geom_point(size = 0.5, alpha=0.3) +
  #geom_point(aes(x = 2, y = 55), color="red") +
  annotate("point", x = c(2, 4.25), y = c(55.43, 81.11), color="red") +
  labs(y = "Time between eruptions (minutes)",
       x = "Duration of eruption (minutes)") +
  theme_classic()
```


```{r}
em_res$mu[[1]]
em_res$mu[[2]]
```


## Multivariate Gaussian mixture model

```{r echo=F, message=F, warning=F, fig.width=3.25, fig.height=3}
data.frame(x = geyser$waiting[2:299], y = geyser$duration[1:298]) |>
  ggplot(aes(x = y, y = x)) +
  geom_point(size = 0.5, alpha=0.3) +
  #geom_point(aes(x = 2, y = 55), color="red") +
  annotate("point", x = c(2, 4.25), y = c(55.43, 81.11), color="red") +
  labs(y = "Time between eruptions (minutes)",
       x = "Duration of eruption (minutes)") +
  stat_ellipse(aes(group = as.factor(round(em_res$posterior[,2]))),
               type="norm", level = 0.5) +
  stat_ellipse(aes(group = as.factor(round(em_res$posterior[,2]))),
               type="norm", level = 0.7) +
  stat_ellipse(aes(group = as.factor(round(em_res$posterior[,2]))),
               type="norm", level = 0.9) +
  theme_classic()
```

## Your turn

Implement the algorithm to fit a Gaussian mixture model:

[https://sta379-s25.github.io/practice_questions/pq_31.html](https://sta379-s25.github.io/practice_questions/pq_31.html)

* Start in class
* Welcome to work with a neighbor
* Solutions are posted on the course website








