---
title: "Lecture 26: Importance Sampling"
author: "Ciaran Evans"
output: beamer_presentation
editor_options: 
  chunk_output_type: console
---

## Last time

$$\theta = \int \limits_0^1 \dfrac{e^{-x}}{1 + x^2} dx$$

**Monte Carlo integration:** write
$$\theta = \int \limits_0^1 \dfrac{e^{-x}}{1 + x^2} dx = \int \limits_0^1 g(x) f(x) dx$$
where $f$ is a density function.

* Sample $X_1,...,X_n \overset{iid}{\sim} f$
* **Monte Carlo estimate:** $\widehat{\theta} = \frac{1}{n} \sum \limits_{i=1}^n g(X_i)$

## Last time

$$\theta = \int \limits_0^1 \dfrac{e^{-x}}{1 + x^2} dx = \int \limits_0^1 g(x) f(x) dx$$

A couple different options:

* $f_1(x) = 1$, $g_1(x) = \dfrac{e^{-x}}{1 + x^2}$

\bigskip

* $f_2(x) = \dfrac{4}{\pi(1 + x^2)}$, $g_2(x) = \frac{\pi}{4} e^{-x}$

\bigskip

**Question:** How would we sample from the distribution with density $f_2$?

## Inverse transform method

pdf: $f_2(x) = \dfrac{4}{\pi(1 + x^2)} \hspace{1cm}$ cdf: $F_2(x) = \frac{4}{\pi}\text{atan}(x)$

To sample $X_1,...,X_n \overset{iid}{\sim} f_2$:

* Sample $U_1,...,U_n \overset{iid}{\sim} Uniform(0, 1)$

\bigskip

* $X_i = F_2^{-1}(U_i)$

\vspace{2cm}

## Variance

```{r}
g2 <- function(x){ pi/4 * exp(-x)}

n <- 10; nsim <- 1000
theta_hat2 <- rep(NA, nsim)

for(i in 1:nsim){
  x <- tan(runif(n) * pi/4)
  theta_hat2[i] <- mean(g2(x))
}

var(theta_hat2)
sd(theta_hat2)
```

```{r, include=F}
g <- function(x){exp(-x)/(1 + x^2)}

n <- 10
nsim <- 1000
theta_hat <- rep(NA, nsim)

for(i in 1:nsim){
  x <- runif(n)
  theta_hat[i] <- mean(g(x))
}

var(theta_hat)
sd(theta_hat)
```

## Comparing variance

$$\theta = \int \limits_0^1 \dfrac{e^{-x}}{1 + x^2} dx = \int \limits_0^1 g(x) f(x) dx$$

* $f_1(x) = 1$, $g_1(x) = \dfrac{e^{-x}}{1 + x^2}$: $Var(\widehat{\theta}_1) \approx 0.006$

* $f_2(x) = \dfrac{4}{\pi(1 + x^2)}$, $g_2(x) = \frac{\pi}{4} e^{-x}$: $Var(\widehat{\theta}_2) \approx 0.002$

**Relative efficiency:** $\dfrac{Var(\widehat{\theta}_2)}{Var(\widehat{\theta}_1)} \approx \dfrac{1}{3}$

**Reduction in variance:** $100 \cdot\dfrac{Var(\widehat{\theta}_1) - Var(\widehat{\theta}_2)}{Var(\widehat{\theta}_1)} \approx 68\%$ reduction in variance

## Comparing options

\vspace{-0.5cm}

$$\theta = \int \limits_0^1 h(x) dx = \int \limits_0^1 g(x) f(x) dx \hspace{1cm} h(x) =\dfrac{e^{-x}}{1 + x^2}$$

\vspace{-0.5cm}

* $f_1(x) = 1$, $g_1(x) = \dfrac{e^{-x}}{1 + x^2}$: $Var(\widehat{\theta}_1) \approx 0.006$

\vspace{-0.25cm}

* $f_2(x) = \dfrac{4}{\pi(1 + x^2)}$, $g_2(x) = \frac{\pi}{4} e^{-x}$: $Var(\widehat{\theta}_2) \approx 0.002$

```{r, message=F, echo=F, fig.width=4, fig.height=2.2}
library(tidyverse)

h <- function(x){
  exp(-x)/(1 + x^2)
}

f1 <- function(x){
  rep(1, length(x))
}

f2 <- function(x){
  4/pi * (1 + x^2)^{-1}
}

gridpts <- seq(0, 1, 0.01)
data.frame(x = rep(gridpts, 3), 
           y = c(h(gridpts), f1(gridpts), f2(gridpts)),
           type = rep(c("h(x)", "f1(x)", "f2(x)"), each=length(gridpts))) |>
  ggplot(aes(x = x, y = y, color=type)) +
  geom_line() +
  labs(x = "x", y = "", color="") +
  theme_bw()
```


## Importance sampling

For any density function $f$ supported on the domain of integration $\mathcal{X}$:

$$\theta = \int \limits_{\mathcal{X}} h(x) dx = \int \limits_{\mathcal{X}} \frac{h(x)}{f(x)}f(x) dx$$

* $X_1,...,X_n \overset{iid}{\sim} f$

\bigskip

* $\widehat{\theta} = \frac{1}{n} \sum \limits_{i=1}^n \dfrac{h(X_i)}{f(X_i)}$

\bigskip

$$\mathbb{E}[\widehat{\theta}] = \theta \hspace{1cm} Var(\widehat{\theta}) = \frac{1}{n} Var \left( \frac{h(X)}{f(X)} \right)$$

**Question:** What choice of $f$ would **minimize** this variance?

## Importance sampling

$$\theta = \int \limits_{\mathcal{X}} h(x) dx = \int \limits_{\mathcal{X}} \frac{h(x)}{f(x)}f(x) dx$$

* $Var(\widehat{\theta})$ is smaller if $f(x)$ is "similar to" $h(x)$

\bigskip

* If $f(x) = c \cdot h(x)$ for all $x$ and some constant $c$, then $Var \left( \dfrac{h(X)}{f(X)} \right) = Var \left( \dfrac{1}{c} \right) = 0$

\bigskip

* $Var(\widehat{\theta})$ is minimized if $f(x) = \dfrac{|h(x)|}{\int |h(t)| dt}$

\bigskip

* But, if we can't integrate $h(x)$ anyway, then it isn't actually possible to use this $f(x)$

## Your turn

Experiment with using different densities for Monte Carlo integration with another integral:

[https://sta379-s25.github.io/practice_questions/pq_26.html](https://sta379-s25.github.io/practice_questions/pq_26.html)

* Start in class
* You are welcome and encouraged to work with your neighbors
* Solutions posted on course website
