---
title: "Lecture 6: Generating random variables -- transformations"
author: "Ciaran Evans"
output: beamer_presentation
---

## Previously

Methods to generate $U \sim Uniform(0, 1)$:

* Linear congruential generator
* Mersenne twister 
* lots of other variants and alternatives

Now we want to generate random variables with *other* distributions.

**Question:** If I have a uniform random variable, how can I get other random variables?

## Inverse transform method


## Example

Suppose we want to generate $X \sim Exponential(\theta)$

\vspace{5cm}

## Example

Generating $X \sim Exponential(1)$:

```{r}
# generate 1000 samples
u <- runif(1000)
x <- -log(u)
```

**Question:** How can I check that the samples match the desired $Exponential(1)$ distribution?

\vspace{3cm}

## Example

Generating $X \sim Exponential(1)$:

```{r}
# generate 1000 samples
u <- runif(1000)
x <- -log(u)
```

```{r, echo=F, message=F, fig.width=6, fig.height=4}
gridpts <- seq(0, 7, 0.01)
plot(gridpts, ecdf(x)(gridpts), type="l", lwd=1.5, xlab="x", ylab="cdf")
lines(gridpts, pexp(gridpts), col="orange", lwd=1.5)
legend(3,0.5,c("Empirical cdf X","Exp(1) cdf"),lty=c(1,1),col=c("black","orange"))
```

## Discrete case

Suppose that we want to generate $X \sim Bernoulli(p)$

\vspace{5cm}

## Generating a Normal random variable

Suppose we want to simulate $X \sim N(0, 1)$

$$f_X(x) = \frac{1}{\sqrt{2 \pi}} \exp \left\lbrace -\frac{1}{2} x^2 \right\rbrace$$

$$F_X(t) = \ ?$$

\vspace{3cm}

## Box-Muller Transformation

## Box-Muller in practice

```{r}
# generate 1000 samples
u1 <- runif(1000)
u2 <- runif(1000)
x1 <- sqrt(-2*log(u1)) * cos(2*pi*u2)
x2 <- sqrt(-2*log(u1)) * sin(2*pi*u2)
```

```{r, echo=F, message=F, fig.width=6, fig.height=4}
gridpts <- seq(-3, 3, 0.01)
plot(gridpts, ecdf(x1)(gridpts), type="l", lwd=1.5, xlab="x", ylab="cdf")
lines(gridpts, pnorm(gridpts), col="orange", lwd=1.5)
legend(0.5,0.5,c("Empirical cdf X1","N(0, 1) cdf"),lty=c(1,1),col=c("black","orange"))
```

## Box-Muller in practice


```{r, echo=F, message=F, fig.width=3, fig.height=3}
library(tidyverse)
data.frame(x1, x2) |>
  ggplot(aes(x = x1, y = x2)) +
  geom_point(size=0.5) +
  theme_bw()
```


## Other Normals

Suppose that $Z \sim N(0, 1)$. How do I get $X \sim N(\mu, \sigma^2)$?

\vspace{4cm}

## A few other transformations

* If $X \sim N(\mu, \sigma^2)$, then $e^X \sim Lognormal(\mu, \sigma^2)$

* If $Z_1,...,Z_k \overset{iid}{\sim} N(0, 1)$, then 

$$\sum \limits_{i=1}^k Z_i^2 \sim \ ?$$

* If $V_1 \sim \chi^2_{d_1}$ and $V_2 \sim \chi^2_{d_2}$ are independent, then 

$$\frac{V_1/d_1}{V_2/d_2} \sim \ ?$$

## Your turn

Practice questions on the course website:

[https://sta379-s25.github.io/practice_questions/pq_6.html](https://sta379-s25.github.io/practice_questions/pq_6.html)

* Practice with inverse transform method
* Start in class. You are welcome to work with others
* Practice questions are to help you practice. They are not submitted and not graded
* Solutions are posted on the course website
