---
title: "Lecture 29: Introducing the EM algorithm"
author: "Ciaran Evans"
output: beamer_presentation
editor_options: 
  chunk_output_type: console
---

## Plan for next week

* Monday: continue EM algorithm

\bigskip

* Wednesday and Friday: project work days

\bigskip

* Extra office hours on Tuesday, Wednesday, and Thursday

## Motivation: penguins data

Data on 276 penguins (Adelie or Gentoo) on three different islands (Torgersen, Biscoe, Dream) near Antartica. Variables include

* species
* island
* characteristics like bill length, flipper length, etc.

## Motivation: penguins data

```{r, fig.width=3.5, fig.height=2.5, message=F, warning=F, echo=F}
library(tidyverse)
library(palmerpenguins)

penguins |>
  filter(species != "Chinstrap") |>
  ggplot(aes(x = flipper_length_mm)) +
  geom_histogram(fill="transparent", color="black") +
  labs(x = "Flipper length (mm)") +
  theme_classic()
```

**Question:** What do you notice about the distribution of flipper length? Why might this be the case?

## Motivation: penguins data


```{r, fig.width=4, fig.height=2, message=F, warning=F, echo=F}
penguins |>
  filter(species != "Chinstrap") |>
  ggplot(aes(x = flipper_length_mm)) +
  geom_histogram(fill="transparent", color="black") +
  labs(x = "Flipper length (mm)") +
  theme_classic() +
  facet_wrap(~species)
```

**Question:** How could I model the distribution of flipper length in each group? What parameters would I estimate?

\vspace{3cm}

## Motivation: penguins data


```{r, fig.width=4, fig.height=3, message=F, warning=F, echo=F}
mean_adelie <- penguins |>
  filter(species == "Adelie") |>
  pull(flipper_length_mm) |>
  mean(na.rm=T)

sd_adelie <- penguins |>
  filter(species == "Adelie") |>
  pull(flipper_length_mm) |>
  sd(na.rm=T)

mean_gentoo <- penguins |>
  filter(species == "Gentoo") |>
  pull(flipper_length_mm) |>
  mean(na.rm=T)

sd_gentoo <- penguins |>
  filter(species == "Gentoo") |>
  pull(flipper_length_mm) |>
  sd(na.rm=T)

prop_adelie = sum(penguins$species == "Adelie")/sum(penguins$species != "Chinstrap")

penguins |>
  filter(species != "Chinstrap") |>
  ggplot(aes(x = flipper_length_mm)) +
  geom_histogram(fill="transparent", color="black", aes(y = after_stat(density))) +
  labs(x = "Flipper length (mm)") +
  theme_classic() +
  stat_function(fun = function(x){prop_adelie*dnorm(x, mean_adelie, sd_adelie)},
                color = "blue") +
  stat_function(fun = function(x){(1 - prop_adelie)*dnorm(x, mean_gentoo, sd_gentoo)},
                color = "red")
```

## Writing down a model


```{r, fig.width=3, fig.height=2, message=F, warning=F, echo=F}

penguins |>
  filter(species != "Chinstrap") |>
  ggplot(aes(x = flipper_length_mm)) +
  geom_histogram(fill="transparent", color="black", aes(y = after_stat(density))) +
  labs(x = "Flipper length (mm)") +
  theme_classic() +
  stat_function(fun = function(x){prop_adelie*dnorm(x, mean_adelie, sd_adelie)},
                color = "blue") +
  stat_function(fun = function(x){(1 - prop_adelie)*dnorm(x, mean_gentoo, sd_gentoo)},
                color = "red")
```


\vspace{3cm}


## Time between Old Faithful geyser eruptions

```{r, fig.width=3.5, fig.height=2.5, message=F, warning=F, echo=F}
library(MASS)

geyser |>
  ggplot(aes(x = waiting)) +
  geom_histogram(fill="transparent", color="black") +
  labs(x = "Time between eruptions (minutes)") +
  theme_classic()
```

**Question:** What do you notice about the distribution of waiting times? Why might this be the case?

## Time between Old Faithful geyser eruptions

```{r, fig.width=3.5, fig.height=2.5, message=F, warning=F, echo=F}
geyser |>
  ggplot(aes(x = waiting)) +
  geom_histogram(fill="transparent", color="black") +
  labs(x = "Time between eruptions (minutes)") +
  theme_classic()
```

**Question:** It seems like there are two groups here, but we don't know what they are. What should we do to estimate both the groups and their distributions?

## Time between Old Faithful geyser eruptions


```{r, fig.width=3, fig.height=2, message=F, warning=F, echo=F}
geyser |>
  ggplot(aes(x = waiting)) +
  geom_histogram(fill="transparent", color="black") +
  labs(x = "Time between eruptions (minutes)") +
  theme_classic()
```

**Model:**

\vspace{3cm}

## Gaussian mixture model

* Observe data $X_1,...,X_n$

\bigskip

* Assume each observation $i$ comes from one of $k$ groups. Let $Z_i \in \{1,...,k\}$ denote the group assignment 
    * The group $Z$ is an unobserved (**latent**) variable

**Model:**

\vspace{5cm}

## Estimating model parameters: EM algorithm

The **EM algorithm** allows us to estimate both the unknown group assignments, *and* the parameters for each group's distribution (we will discuss the details later). In R:

```{r, eval=F}
library(mixtools)

normalmixEM(geyser$waiting, lambda = c(0.5, 0.5), k=2)
```

* `normalmixEM`: function for estimating parameters in a mixture of normal distributions
* `lambda`: initial guess at the proportion of data in each group
* `k`: number of groups

## Estimating model parameters: EM algorithm

```{r, include=F}
set.seed(3)
```


```{r, message=F}
library(mixtools)

em_res <- normalmixEM(geyser$waiting, lambda = c(0.5, 0.5), 
                      k=2)

em_res$lambda
em_res$mu
em_res$sigma
```

## Fitted parameters

```{r, fig.width=3.5, fig.height=2.5, message=F, warning=F, echo=F}
geyser |>
  ggplot(aes(x = waiting)) +
  geom_histogram(fill="transparent", color="black", aes(y = after_stat(density))) +
  labs(x = "Time between eruptions (minutes)") +
  theme_classic() +
  stat_function(fun = function(x){0.308*dnorm(x, 54.203, 4.951)},
                color = "blue") +
  stat_function(fun = function(x){(1 - 0.308)*dnorm(x, 80.360, 7.508)},
                color = "red")
```

* Estimated proportion of data in each group: 0.308, 0.692
* Estimated group means: $\widehat{\mu}_1 = 54.203$, $\widehat{\mu}_2 = 80.360$
* Estimated group sd: $\widehat{\sigma}_1 = 4.951$, $\widehat{\sigma}_2 = 7.508$


## Your turn

Simulate data from a Gaussian mixture and explore parameter estimation:

[https://sta379-s25.github.io/practice_questions/pq_29.html](https://sta379-s25.github.io/practice_questions/pq_29.html)

* Start in class
* Welcome to work with a neighbor
* Solutions will be posted later on the course website




