---
title: "Lecture 17: Newton's method"
author: "Ciaran Evans"
output: beamer_presentation
editor_options: 
  chunk_output_type: console
---

## Motivation

:::: {.columns}

::: {.column width="50%"}

$$f(x) = \frac{x^4}{4} - \sin(x) + 7$$

```{r, echo=F, fig.align='center', fig.width=2.8, fig.height=3}
fx <- function(x){
  x^4/4 - sin(x) + 7
}
gridpts <- seq(-3, 3, 0.01)
plot(gridpts, fx(gridpts), type="l",
     xlab = "x", ylab = "f(x)", cex.lab=0.75, cex.axis=0.75)
```
:::

::: {.column width="50%"}

$$f'(x) = x^3 - \cos(x) $$

```{r, echo=F, fig.align='center', fig.width=2.8, fig.height=3}
fp <- function(x){
  x^3 - cos(x)
}
gridpts <- seq(-3, 3, 0.01)
plot(gridpts, fp(gridpts), type="l",
     xlab = "x", ylab = "f'(x)", cex.lab=0.75, cex.axis=0.75)
```
:::

::::

Want to minimize $f(x)$, i.e. solve $f'(x) = 0$. Challenge: no closed form.

**Idea:** Approximate $f'$ with something *easy* to solve

## Taylor expansion

First-order Taylor expansion of function $g(x)$ around point $a$:

$$g(x) \approx g(a) + (x - a)g'(a)$$

**Question:** What is required of $x$ for this approximation to be reasonable?

\vspace{4cm}

## Taylor expansion

First-order Taylor expansion of function $g(x)$ around point $a$:

$$g(x) \approx g(a) + (x - a)g'(a)$$

* Let $x^*$ be the location of the true minimum
* Let $x^{(k)}$ be our current guess for $x^*$
* Taylor expansion:

$$f'(x^*) \approx \hspace{4cm}$$

\vspace{4cm}

## Newton's method (in one dimension)

Iterative updates:

$$x^{(k+1)} = x^{(k)} - (f''(x^{(k)}))^{-1} f'(x^{(k)})$$

**Question:** How does this compare with gradient descent?

\vspace{5cm}

## Newton's method (in one dimension)

Iterative updates:

$$x^{(k+1)} = x^{(k)} - (f''(x^{(k)}))^{-1} f'(x^{(k)})$$

More generally:

$$x^{(k+1)} = x^{(k)} - \alpha_k (f''(x^{(k)}))^{-1} f'(x^{(k)})$$

* $\alpha_k = 1$ is the "natural" value
* Only use a different value of $\alpha_k$ when $\alpha_k = 1$ doesn't work

## Illustration

```{r, echo=F, fig.width=4, fig.height=3.5}
fpp <- function(x){
  3*x^2 + sin(x)
}

fp_taylor <- function(x, a){
  fp(a) + (x-a)*fpp(a)
}
plot(gridpts, fp(gridpts), type="l",
     xlab = "x", ylab = "f'(x)", cex.lab=0.75, cex.axis=0.75)
xk <- -2
points(gridpts, fp_taylor(gridpts, xk), type="l", col="orange")
```

* $x^{(0)} = -2$

## Illustration

```{r, echo=F, fig.width=4, fig.height=3.5}
fpp <- function(x){
  3*x^2 + sin(x)
}

fp_taylor <- function(x, a){
  fp(a) + (x-a)*fpp(a)
}
plot(gridpts, fp(gridpts), type="l",
     xlab = "x", ylab = "f'(x)", cex.lab=0.75, cex.axis=0.75)
xk <- -2
points(gridpts, fp_taylor(gridpts, xk), type="l", col="orange")
abline(h=0, lty=2)
```

* $x^{(0)} = -2$
* $x^{(1)}$ solves $f'(x^{(0)}) + (x^{(1)} - x^{(0)})f''(x^{(0)}) = 0$

## Illustration

```{r, echo=F, fig.width=4, fig.height=3.5}
fpp <- function(x){
  3*x^2 + sin(x)
}

fp_taylor <- function(x, a){
  fp(a) + (x-a)*fpp(a)
}
plot(gridpts, fp(gridpts), type="l",
     xlab = "x", ylab = "f'(x)", cex.lab=0.75, cex.axis=0.75)
xk <- -2
points(gridpts, fp_taylor(gridpts, xk), type="l", col="orange")
abline(h=0, lty=2)

xk <- xk - fpp(xk)^{-1}*fp(xk)
abline(v = xk, lty=2)
```

* $x^{(1)}$ solves $f'(x^{(0)}) + (x^{(1)} - x^{(0)})f''(x^{(0)}) = 0$
* $x^{(1)} = -1.316$

## Illustration

```{r, echo=F, fig.width=4, fig.height=3.5}
fpp <- function(x){
  3*x^2 + sin(x)
}

fp_taylor <- function(x, a){
  fp(a) + (x-a)*fpp(a)
}
plot(gridpts, fp(gridpts), type="l",
     xlab = "x", ylab = "f'(x)", cex.lab=0.75, cex.axis=0.75)
xk <- -1.316197
points(gridpts, fp_taylor(gridpts, xk), type="l", col="orange")
abline(h=0, lty=2)

xk <- xk - fpp(xk)^{-1}*fp(xk)
abline(v = xk, lty=2)
```

* $x^{(1)} = -1.316$
* $x^{(2)}$ solves $f'(x^{(1)}) + (x^{(2)} - x^{(1)})f''(x^{(1)}) = 0$

## Extension to multiple variables

Example: $f(x,y) = \dfrac{x^4}{4} - \sin(x) + (x + y)^2$

$$\nabla f(x, y) = \begin{pmatrix} \dfrac{\partial f}{\partial x} \\ \dfrac{\partial f}{\partial y} \end{pmatrix} =  \begin{pmatrix} x^3 - \cos(x) + 2(x + y) \\ 2(x + y) \end{pmatrix}$$

**Question:** What is the *second* derivative of $f$?

\vspace{5cm}

## Extension to multiple variables

Example: $f(x,y) = \dfrac{x^4}{4} - \sin(x) + (x + y)^2$

$$\nabla f(x, y) = \begin{pmatrix} \dfrac{\partial f}{\partial x} \\ \dfrac{\partial f}{\partial y} \end{pmatrix} =  \begin{pmatrix} x^3 - \cos(x) + 2(x + y) \\ 2(x + y) \end{pmatrix}$$

**Hessian:** ${\bf H}_f(x, y) = \begin{bmatrix} \dfrac{\partial^2 f}{\partial x^2} & \dfrac{\partial^2 f}{\partial x \partial y} \\ \dfrac{\partial^2 f}{\partial y \partial x} & \dfrac{\partial^2 f}{\partial y^2} \end{bmatrix}$

\vspace{5cm}

## Newton's method in multiple dimensions

* ${\bf x} = (x_1,...,x_d)^T \in \mathbb{R}^d$
* $f({\bf x}) \in \mathbb{R}$

$$\nabla f({\bf x}) = \begin{pmatrix} \frac{\partial f}{\partial x_1} \\ \vdots \\ \frac{\partial f}{\partial x_d} \end{pmatrix} \hspace{1cm} {\bf H}_f({\bf x}) = \begin{bmatrix} \frac{\partial^2 f}{\partial x_1^2} & \frac{\partial^2 f}{\partial x_1 \partial x_2} & \cdots  & \frac{\partial^2 f}{\partial x_1 \partial x_d} \\ \frac{\partial^2 f}{\partial x_2 \partial x_1} & \frac{\partial^2 f}{\partial x_2^2} & \cdots & \frac{\partial^2 f}{\partial x_2 \partial x_d} \\ \vdots & \vdots& \cdots & \vdots \\ \frac{\partial^2 f}{\partial x_d \partial x_1} & \frac{\partial^2 f}{\partial x_d \partial x_2} & \cdots & \frac{\partial^2 f}{\partial x_d^2} \end{bmatrix}$$

**Newton's method:**

$${\bf x}^{(k+1)} = {\bf x}^{(k)} - \alpha_k({\bf H}_f({\bf x}^{(k)}))^{-1} \nabla f({\bf x}^{(k)})$$

```{r, include=F}
f <- function(v){
  x <- v[1]
  y <- v[2]
  x^4/4 - sin(x) + (x + y)^2
}

fp <- function(v){
  x <- v[1]
  y <- v[2]
  c(x^3 - cos(x) + 2*(x + y),
    2*(x + y))
}

Hf <- function(v){
  x <- v[1]
  y <- v[2]
  matrix(c(3*x^2 + sin(x) + 2, 2,
           2, 2), nrow=2)
}

xk <- c(-2, 0)
xk <- xk - c(solve(Hf(xk)) %*% fp(xk))
xk
```

```{r, include=F}
x <- seq(-2, 2, by=0.005)
y <- seq(-2, 2, by=0.005)
z <- matrix(nrow=length(x), ncol=length(y))
for(i in 1:nrow(z)){
  for(j in 1:ncol(z)){
    z[i,j] <- f(c(x[i], y[j]))
  }
}

contour(x, y, z)

xk <- c(-2, 0)
points(xk[1], xk[2], pch=16, col="blue")
```

## Your turn

* Newton's method practice questions on the course website:

[https://sta379-s25.github.io/practice_questions/pq_17.html](https://sta379-s25.github.io/practice_questions/pq_17.html)

* If done early, go back to momentum practice questions from Monday:

[https://sta379-s25.github.io/practice_questions/pq_16.html](https://sta379-s25.github.io/practice_questions/pq_16.html)


