---
title: "Lecture 21: Gaussian quadrature"
author: "Ciaran Evans"
output: beamer_presentation
editor_options: 
  chunk_output_type: console
---

## Gaussian quadrature

**General result:** If $f(x) = c_0 + c_1 x + c_2x^2 + \cdots + c_{2n-1} x^{2n-1}$, then there exist **nodes** $x_1,...,x_n$ and **weights** $w_1,...,x_n$ such that

$$\int \limits_{-1}^1 f(x) dx = \sum \limits_{i=1}^n w_i f(x_i)$$

**n-node Gaussian quadrature rule:** For general function $f$,

$$\int \limits_{-1}^1 f(x) dx \approx \sum \limits_{i=1}^n w_i f(x_i)$$

## Activity, Part 1

Class activity to help motivate Gaussian quadrature:

[https://sta379-s25.github.io/practice_questions/pq_21.html](https://sta379-s25.github.io/practice_questions/pq_21.html)

* Work with your neighbors on Part 1 of the activity
* In a bit, we will discuss key points as a class

## Activity, Part 1

$$L_{n,i}(x) = \prod \limits_{k : k \neq i} \frac{(x - x_k)}{(x_i - x_k)}$$

\vspace{1cm}

$$q(x) = \sum \limits_{i=1}^n y_i L_{n,i}(x)$$

$$q(x_i) = \hspace{3cm}$$

\vspace{2cm}

## Activity, Part 1

Original: \textcolor{red}{$f(x) = 10(x^7 - 1.6225x^5 +0.79875x^3 - 0.113906x)$}

Polynomial interpolation: $q(x)$, with $n=4$ points

```{r, echo=F, fig.width=4, fig.height=3.5}
# calculate q at a single point
# x: point to evaluate q(x)
# xi: the points x1,...,xn
# yi: the points y1,...,yn
q_helper <- function(x, xi, yi){
  lp <- sapply(1:length(xi), 
               function(i){prod((x - xi[-i])/(xi[i] - xi[-i]))})
  sum(yi*lp)
}

# calculate q at a vector of new points
# x: point to evaluate q(x)
# xi: the points x1,...,xn
# yi: the points y1,...,yn
q <- function(x, xi, yi){
  sapply(x, function(t){q_helper(t, xi, yi)})
}

f <- function(x){
  10*(x^7 -1.6225*x^5 +0.79875*x^3 - 0.113906*x)
}

n <- 4
xi <- seq(-1, 1, length.out=n)
yi <- f(xi)

plot(xi, yi, pch=16, xlab="x", ylab="y", cex.lab=0.75, cex.axis=0.75)

x <- seq(-1, 1, 0.01)
lines(x, q(x, xi, yi))
lines(x, f(x), col="red")
```

## Activity, Part 1

Polynomial interpolation: $q(x)$, with $n=4$ points

```{r, echo=F, fig.width=4, fig.height=3.5}
# calculate q at a single point
# x: point to evaluate q(x)
# xi: the points x1,...,xn
# yi: the points y1,...,yn
q_helper <- function(x, xi, yi){
  lp <- sapply(1:length(xi), 
               function(i){prod((x - xi[-i])/(xi[i] - xi[-i]))})
  sum(yi*lp)
}

# calculate q at a vector of new points
# x: point to evaluate q(x)
# xi: the points x1,...,xn
# yi: the points y1,...,yn
q <- function(x, xi, yi){
  sapply(x, function(t){q_helper(t, xi, yi)})
}

f <- function(x){
  10*(x^7 -1.6225*x^5 +0.79875*x^3 - 0.113906*x)
}

n <- 4
xi <- seq(-1, 1, length.out=n)
yi <- f(xi)

plot(xi, yi, pch=16, xlab="x", ylab="y", cex.lab=0.75, cex.axis=0.75)

x <- seq(-1, 1, 0.01)
lines(x, q(x, xi, yi))
lines(x, f(x), col="red")
```

**Question:** What happens as I change the number of points $n$ used for $q(x)$?

## Activity, Part 1

Polynomial interpolation: $q(x)$, with $n=6$ points

```{r, echo=F, fig.width=4, fig.height=3.5}
# calculate q at a single point
# x: point to evaluate q(x)
# xi: the points x1,...,xn
# yi: the points y1,...,yn
q_helper <- function(x, xi, yi){
  lp <- sapply(1:length(xi), 
               function(i){prod((x - xi[-i])/(xi[i] - xi[-i]))})
  sum(yi*lp)
}

# calculate q at a vector of new points
# x: point to evaluate q(x)
# xi: the points x1,...,xn
# yi: the points y1,...,yn
q <- function(x, xi, yi){
  sapply(x, function(t){q_helper(t, xi, yi)})
}

f <- function(x){
  10*(x^7 -1.6225*x^5 +0.79875*x^3 - 0.113906*x)
}

n <- 6
xi <- seq(-1, 1, length.out=n)
yi <- f(xi)

plot(xi, yi, pch=16, xlab="x", ylab="y", cex.lab=0.75, cex.axis=0.75)

x <- seq(-1, 1, 0.01)
lines(x, q(x, xi, yi))
lines(x, f(x), col="red")
```

## Activity, Part 1

Polynomial interpolation: $q(x)$, with $n=8$ points

```{r, echo=F, fig.width=4, fig.height=3.5}
# calculate q at a single point
# x: point to evaluate q(x)
# xi: the points x1,...,xn
# yi: the points y1,...,yn
q_helper <- function(x, xi, yi){
  lp <- sapply(1:length(xi), 
               function(i){prod((x - xi[-i])/(xi[i] - xi[-i]))})
  sum(yi*lp)
}

# calculate q at a vector of new points
# x: point to evaluate q(x)
# xi: the points x1,...,xn
# yi: the points y1,...,yn
q <- function(x, xi, yi){
  sapply(x, function(t){q_helper(t, xi, yi)})
}

f <- function(x){
  10*(x^7 -1.6225*x^5 +0.79875*x^3 - 0.113906*x)
}

n <- 8
xi <- seq(-1, 1, length.out=n)
yi <- f(xi)

plot(xi, yi, pch=16, xlab="x", ylab="y", cex.lab=0.75, cex.axis=0.75)

x <- seq(-1, 1, 0.01)
lines(x, q(x, xi, yi))
lines(x, f(x), col="red")
```

## Polynomial interpolation

Let $f$ be a function on $(-1, 1)$ we wish to approximate, and let $x_1,...,x_n$ be $n$ distinct points in $(-1, 1)$.

**Interpolating polynomial:** $q(x) = \sum \limits_{i=1}^n f(x_i) L_{n,i}(x)$

* $q(x_i) = f(x_i)$ for $i=1,...,n$ (**interpolation**)
* If $f$ is a polynomial of degree $\leq n-1$, then $q(x) = f(x)$ for **all** $x$

Integration: $\int \limits_{-1}^1 f(x) dx \approx \int \limits_{-1}^1 q(x) dx$

\vspace{4cm}

## Summary so far

## Summary so far

To approximate $\int \limits_{-1}^1 f(x) dx$:

1. Choose $n$ points $x_1,...,x_n$ in $(-1, 1)$
2. Construct the interpolating polynomial: $q(x) = \sum \limits_{i=1}^n f(x_i) L_{n,i}(x)$
3. Integrate $q$:
$$\int \limits_{-1}^1 q(x) dx = \sum \limits_{i=1}^n w_i f(x_i) \hspace{1cm} w_i = \int\limits_{-1}^1 L_{n,i}(x) dx$$
4. Approximate the integral of $f$:
$$\int \limits_{-1}^1 f(x)dx \approx \int \limits_{-1}^1 q(x) dx = \sum \limits_{i=1}^n w_i f(x_i)$$

**Next time:** Which points $x_1,...,x_n$ do we use??


## Activity, Part 2

Verify calculation of the weights $w_i$:

[https://sta379-s25.github.io/practice_questions/pq_21.html](https://sta379-s25.github.io/practice_questions/pq_21.html)

* Work with your neighbors on Part 2 of the activity
