---
title: "Practice Questions: Compass search"
format: html
---

## Logistic regression

We have data on medical school admissions for 55 students from a liberal arts college. The data include the variables

* `Acceptance`: whether the student was accepted (1) or not accepted (0) to med school
* `GPA`: the student's college GPA (4-point scale)

We would like to build a model to predict acceptance using college GPA. Because acceptance is a binary variable, a **logistic regression** model is appropriate.

Let $p_i = P(\text{Acceptance}_i = 1)$ be the probability of acceptance for the $i$th individual in the data. Then, the logistic regression model is

$$\log \left(\frac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 \text{GPA}_i$$

Fitting the model in R:

```{r}
library(Stat2Data)
data("MedGPA")

glm(Acceptance ~ GPA, family = binomial, data = MedGPA)
```

### Optimization

For linear regression, we minimized the RSS. What about logistic regression? 

It turns out that the logistic regression model here estimates $\beta_0$ and $\beta_1$ by minimizing the following function:

$$f(\beta_0, \beta_1) = - \sum \limits_{i=1}^n \left \lbrace \text{Acceptance}_i(\beta_0 + \beta_1 \text{GPA}_i) - \log(1 + e^{\beta_0 + \beta_1 \text{GPA}_i}) \right\rbrace$$

The function below calculates this for an input vector $\beta = (\beta_0, \beta_1)$:

```{r}
logistic_loss <- function(beta){
  odds <- exp(beta[1] + beta[2]*MedGPA$GPA)
  p <- odds/(1 + odds)
  -1*sum(dbinom(MedGPA$Acceptance, 1, p, log=T))
}
```


## Compass search

Begin with initial values $(\beta_0^{(0)}, \beta_1^{(0)}) = (0, 0)$ and initial step size $\Delta_0 = 1$.

1. Find the test points around the initial guess, and calculate $f(\beta_0, \beta_1)$ at each of the test points. Will the compass search algorithm move to one of the new points, or shrink the step size?

2. Update the test points based on question 1, and recalculate $f(\beta_0, \beta_1)$ at each. Will the compass search algorithm move to one of the new points, or shrink the step size?

3. Write code to repeat the compass search until the step size $\Delta_k < 1 \times 10^{-6}$. Your final point should be very close to the estimates from the `glm` function in R.

4. How many steps did you take in question 3?

