---
title: "Warmup: antithetic sampling"
format: html
---

# Monte Carlo integration

Suppose we wish to approximate the integral

$$\theta = \int \limits_0^1 e^x dx = \mathbb{E}[g(U)]$$

where $U \sim Uniform(0, 1)$ and $g(x) = e^x$.

## Simple Monte Carlo estimate

The simple Monte Carlo estimate does the following:

* Sample $U_1,...,U_n \overset{iid}{\sim} Uniform(0, 1)$
* $\widehat{\theta}_{MC} = \frac{1}{n} \sum \limits_{i=1}^n g(U_i)$

Here is code to approximate the variance of the simple MC estimate with $n = 1000$:

```{r}
g <- function(x){
  exp(x)
}

n <- 1000
nsim <- 1000

theta_hat_mc <- rep(NA, nsim)
for(i in 1:nsim){
  u <- runif(n)
  theta_hat_mc[i] <- mean(g(u))
}

var(theta_hat_mc)
```

## Antithetic sampling

One method for reducing the variance of a Monte Carlo estimate is with **antithetic sampling**:

* Sample $U_1,...,U_{n/2} \sim Uniform(0, 1)$
* $\widehat{\theta}_{AS} = \frac{1}{n} \sum \limits_{i=1}^{n/2} (g(U_i) + g(1 - U_i))$

## Questions

1. Using R, approximate the variance of the antithetic estimator $\widehat{\theta}_{AS}$ with $n = 1000$, and compute the percent reduction in variance compared to the simple estimator $\widehat{\theta}_{MC}$.

2. Using R, approximate $\rho = Cor(g(U), g(1 - U))$ and compare to the percent reduction in variance from question 1.

```{r, include=F}
theta_hat_as <- rep(NA, nsim)
for(i in 1:nsim){
  u <- runif(n/2)
  theta_hat_as[i] <- sum(g(u) + g(1-u))/n
}

var(theta_hat_as)

(var(theta_hat_mc) - var(theta_hat_as))/var(theta_hat_mc)

u <- runif(10000)
cor(g(u), g(1-u))
```

