---
title: "Review practice questions: Functions and scoping"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

# Function recap

Here is a simple function to calculate the absolute value (redundant since R already has the `abs` function, but we'll still use it as a toy example):

```{r, echo=T}
my_abs <- function(x){
  return(ifelse(x >= 0, x, -1*x))
}

my_abs(-3)
my_abs(c(-2, 5))
```

Functions in R (or other programming languages) behave very similarly to mathematical functions. Each function performs a certain task on the inputs, and produces some result. In this example:
* the function is called `my_abs`
* the input to the function is a vector `x`
* and the output is another vector, containing the absolute value of the entries of `x`. 

Note that by using the `ifelse` function here, we have vectorized the `my_abs` function.

## Ordering and arguments

Order of arguments doesn't matter if you name the arguments when calling the function. If you don't name the arguments when calling a function, R assumes you passed them in the order of the function definition:

```{r, echo=T}
my_power <- function(x, y){
  return(x^y)
}
```

```{r, echo=T}
my_power(x = 2, y = 3)
my_power(y = 3, x = 2)
my_power(2, 3)
my_power(3, 2)
```

All arguments which do not have a default value must be specified when calling the function. For example, the following code produces errors:

```{r, error=T}
my_power(3)
```


## Function defaults

Default arguments can be specified when the function is defined:

```{r, echo=T}
my_power <- function(x, y=2){
  return(x^y)
}

my_power(3)
```

If you don't specify the value of that argument, the default is used. You can, of course, choose to use a different value:

```{r}
my_power(2, 3)
```



# Questions

## Writing a function to calculate the mean

The *mean* of numbers $x_1,...,x_n$ is their average: 

$$\frac{x_1 + x_2 + \cdots + x_n}{n}$$
In R, we can calculate the mean with the `mean` function. For example:

```{r}
mean(c(1,2,3))
```

:::{.question}
#### Question 1

Write your own mean function, called `my_mean`, that calculates the mean of a vector. The input to your function should be a vector, and the output should be the mean of the values in that vector. You may *not* use the built-in `mean` function in R, but you *may* use the `sum` and `length` functions.
:::

:::{.question}
#### Question 2

Check that your function works by comparing the output with R's `mean` function for several different input vectors.
:::

## Weighted means

A *weighted* mean is similar to the usual average, but now we add a *weight* to each value. The observations with greater weights contribute more to the weighted mean.

The weighted mean of $x_1,...,x_n$ with weights $w_1,...,w_n$ is 

$$\frac{w_1 x_1 + w_2 x_2 + \cdots + w_n x_n}{w_1 + w_2 + \cdots + w_n}$$
The usual arithmetic mean is a special case of the weighted mean with $w_1 = w_2 = \cdots = w_n = 1$.

:::{.question}
#### Question 3
Modify your `my_mean` function from Exercise 1 so that it now takes **two** inputs -- a vector of values `x`, and a vector of weights `w` -- and returns the *weighted* mean of `x` with weights `w`. *Hint*: If `x` and `w` are two vectors of the same length, then `x*w` is a vector created by multiplying each entry of `x` with the corresponding entry of `w`.
:::

:::{.question}
#### Question 4
When we calculate a mean, we usually don't need weights. Modify your `my_mean` function from Exercise 3 so that the *default* weights are all 1.
:::

:::{.question}
#### Question 5

Check that your function from Exercise 4 works by running the following:

```{r, eval=F}
# should be 2
my_mean(x = c(1, 2, 3))

# should be 2
my_mean(x = c(1, 2, 3), w = c(1, 1, 1))

# should be 1.5
my_mean(x = c(1, 2, 3), w = c(1, 1, 0))
```
:::

## Standard deviation

The *sample standard deviation* of numbers $x_1,...,x_n$ is given by 

$$\widehat{\sigma} = \sqrt{\frac{1}{n-1}\sum \limits_{i=1}^n (x_i - \bar{x})^2},$$

where $\bar{x} = \frac{1}{n} \sum \limits_{i=1}^n x_i$.

:::{.question}
#### Question 6

Write a function called `my_sd` which calculates the standard deviation of a vector in R.
:::

## p-norm

The $\ell_p$ norm of a vector $x = (x_1,...,x_k)$ is given by

$$||x||_p = \left( \sum \limits_{i=1}^k |x_i|^p \right)^{1/p}$$

:::{.question}
#### Question 7

Write a function called `p_norm` in R, which takes two inputs: a vector `x`, and `p`, and returns $\ell_p(x)$. Make `p = 2` the default value (this corresponds to the usual Euclidean norm).
:::


## Kurtosis

Suppose we have a sample $X_1,...,X_n$ from some population distribution. We know that the mean describes the "center" of the distribution, the standard deviation is a measure of variability, and skewness describes the shape of the distribution. 

Another quantity we can calculate to describe a distribution is *kurtosis*, which describes how heavy the tails of the distribution are. The *sample kurtosis* is calculated by:

$$\dfrac{\frac{1}{n} \sum \limits_{i=1}^n (X_i - \bar{X})^4}{\left( \frac{1}{n} \sum \limits_{i=1}^n (X_i - \bar{X})^2 \right)^2} \ \ - \ \ 3$$

where $\bar{X}$ is the sample mean.

:::{.question}
#### Question 8

Write a function in R to calculate the sample kurtosis. Your function should take in one argument: a vector `x`.
:::

## Correlation

Suppose we have a sample $(X_1, Y_1),...,(X_n, Y_n)$ of $n$ observations collected on two variables, $X$ and $Y$. The strength of the linear relationship between $X$ and $Y$ is measured by their *correlation*, and the sample correlation is calculated with the following formula:


$$\dfrac{ \sum \limits_{i=1}^n (X_i - \bar{X})(Y_i - \bar{Y})}{\left( \sum \limits_{i=1}^n (X_i - \bar{X})^2 \right)^{1/2} \left( \sum \limits_{i=1}^n (Y_i - \bar{Y})^2 \right)^{1/2}}$$

:::{.question}
#### Question 9

Write a function in R to calculate the sample correlation. Your function should take in two arguments: a vector `x` and a vector `y`.
:::

## Function scoping and output

For each question, predict what will happen when the code is run. Then run the code and check whether your prediction was correct.

10. 

```{r, eval=F}
x <- 10

g02 <- function(x){
  x <- x + 1
  return(x)
}

g02(x)
x + 1
```

11.

```{r, eval=F}
x <- 10

g02 <- function(x){
  x <- x + 1
  return(x)
}

x <- g02(x)
x + 1
```

12. 

```{r, eval=F}
g02 <- function(y){
  y <- y - 1
  return(y)
}

g02(g02(20))
```

13. 

```{r, eval=F}
x <- 10
test_fun <- function(x = 11){
  return(x)
}
test_fun()
x
```

14. 

```{r, eval=F}
x <- 10
test_fun <- function(y = 11){
  return(x + 1)
}
test_fun()
x
```

15. 

```{r, eval=F}
x <- 10
test_fun <- function(y = 11){
  x <- x + 1
  return(x + 1)
}
test_fun()
x
```

16. 

```{r, eval=F}
x <- 10
test_fun <- function(x = 11){
  x <- x + 1
  return(x + 1)
}
test_fun()
x
```

17. 

```{r, eval=F}
x <- 10
test_fun <- function(x = 11){
  x <- x + 1
  return(x + 1)
}
x <- test_fun(x)
x
```

18.

```{r, eval=F}
f1 <- function(x = 1){
  return(x + 1)
}
g1 <- function(x){
  return(f1() + x)
}

f1(g1(3))
```

## Practice with anonymous functions

19. Use `integrate()` and an anonymous function to find the area under the curve for the following functions:

* `y = x^2 - x` for x in $[0, 1]$
* `y = sin(x) + cos(x)` for x in $[-\pi, \pi]$
* `y = exp(x)/x` for x in $[10, 20]$

