---
title: "Review practice questions: lists"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

# Review: vectors and lists

## vectors

A *vector* is a simple way of storing multiple items of the same basic type. For example, here is a vector containing the numbers 2, 1, 4:

```{r}
v <- c(2, 1, 4)
v
```

The `c()` in R means "combine" or "concatenate" the elements of the vector together. The items inside the parentheses are the elements of the vector.

### Length

Every vector has a *length*, which tells us how many elements it contains:

```{r}
length(v)
```

### Indexing

We can access each item in the vector by its *index*. That is, we specify the position of the vector, and get back the element at that position:

```{r}
v[1]

v[2]

v[3]
```

We specify the index inside the square brackets `[...]`. For example, `v[2]` is the second element of the vector `v`.

Vectors can store elements other than numbers. For example, here is a vector containing the letters 'a', 'b', 'c', and 'd':

```{r}
w <- c('a', 'b', 'c', 'd')
w
```

We can combine two vectors together with the `c()` function:

```{r}
c(v, w)
```

### Sequences (`seq`)

Suppose we want to create a vector containing the numbers 0, 0.1, 0.2,...,0.9, 1. In other words, a sequence of numbers between 0 and 1, incremented by steps of 0.1. We can use the `seq` function:

```{r}
seq(from=0, to=1, by=0.1)
```

Here's a sequence from 1 to 10, incremented by steps of 1:

```{r}
seq(from=1, to=10, by=1)
```

When we want to increment by steps of 1, R has a handy shorthand:

```{r}
1:10
```

### Repetitions (`rep`)

Other times, we want to create a vector containing one value repeated many times. We can use the `rep` function. For example, suppose we want to create a vector of eleven 0s:

```{r}
rep(0, 11)
```

## Lists

Vectors can contain numbers, booleans, characters, etc:

```{r, echo=T}
x <- c(0, 1, 2)
x
typeof(x)

x <- c("a", "b", "c")
x
typeof(x)
```

The `typeof` function tells what *type* of object we have. However, basic vectors (called *atomic* vectors) only contain one type:


```{r, echo=T, error=T}
x <- c(0, 1, "a")
x
x[1] + 1
```

*Lists* provide the ability to store multiple types in a single object. For example:

```{r, echo=T}
x <- list(c(0, 1), "a")
x
```

Vectors are indexed with single square brackets `[ ]`. Lists are also indexed with square brackets:

```{r}
x[[1]]

x[[1]][1]

typeof(x[[1]])

x[[2]]

typeof(x[[2]])
```

Here is another example:

```{r, echo=T}
x <- list(c(1, 2), c(3, 4))

x[1]
typeof(x[1])

x[[1]]
typeof(x[[1]])
```

* `x[1]` returns a *list* which contains the first component of `x`
* `x[[1]]` returns the object stored in the first component

Note that most things in R are objects; this includes vectors, lists, data frames, functions, etc. This means that I could, for example, create a *list of functions*:

```{r, echo=T}
x <- list(rexp, rnorm)
x[[1]]
x[[1]](10)
```


### Named lists

We can give the entries of lists names. For example, consider the following:

```{r}
course <- list("number" = "STA379",
               "instructor" = "Evans",
               "room" = "Manchester 229",
               "students" = c("Martin Wainwright", "Michael Jordan", "Regina Liu"))
```

We can then access entries of the list in different ways:

```{r}
course[[1]]
course[["number"]]
course$number
```

In particular, you will see the `$` method of accessing list entries frequently in R. For example, consider the output of a linear mode:

```{r}
x <- rnorm(100)
y <- 0.2*x + rnorm(100)
m1 <- lm(y ~ x)

typeof(m1)
length(m1)
m1$coefficients
```

Here the model output is a list, with named entries like `coefficients`.


# Practice with lists

1. Each of the following lists contains the vector c(2, 7, 9). Determine the correct list indexing syntax to extract that vector from the list (your code should return just the vector, not a list containing the vector).

```{r, eval=F}
x1 <- list(c(2, 7, 9))
x2 <- list(list(c(2, 7, 9)))
x3 <- list(c(2, 7, 9), list("a", "b"))
x4 <- list(c("a", "b"), list(list(c(2, 7, 9))))
```

2. Create a list `x5` such that the vector `c(2, 7, 9)` can be extracted with `x5[[3]][[2]][[1]]`

3. Create a list `x` in R such that:
    * `x[[1]]` returns the function `mean`
    * `x[[2]]` returns the function `sd`
    * `x[[3]][[1]]` returns the vector `c(0, 1, 2)`
    * `x[[3]][[2]]` returns a function which calculates the cube root of a vector
    





